---
title: 'Human RNA differential expression in DIPG/DMG versus normal brain'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2025"
---

This script performs differential expression analysis of human RNA transcripts in DIPG/DMG samples versus normal pons

Load packages

```{r load packages}
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(EnhancedVolcano)
library(rtracklayer)
```

Set directory paths

```{r dir paths}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-rna-differential-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

source(file.path(root_dir, "figures", "theme.R"))
```

Set file paths 

```{r file paths}
gene_cts_file <- file.path(data_dir,
                          "30-992989426-RNA_raw_counts.csv")
```

Wrangle data

```{r wrangle data}
gene_cts <- read_csv(gene_cts_file) %>%
  column_to_rownames("...1")
```

Create sample metaata df

```{r metadata}
metadata_df <- tibble(id = names(gene_cts)) %>%
  dplyr::mutate(group = case_when(
    grepl("left-pons", id) ~ "normal pons",
    TRUE ~ "DIPG_DMG"
  )) %>%
  as.data.frame()
```

Construct DESeqDataSet Object

```{r create dds}
dds <- DESeqDataSetFromMatrix(countData = gene_cts, 
                              colData = metadata_df, 
                              design = ~group)
```

Plot gene expression PCA

```{r pca}
vsdata <- varianceStabilizingTransformation(dds)

pdf(file.path(plot_dir,
              "rna-pca-plot.pdf"),
    width = 5, height = 3)

plotPCA(vsdata, intgroup="group") +
  theme_Publication()

dev.off()
```

Run DESeq2 function

```{r deseq2}
dds <- DESeq(dds)
```

Write results to variable

```{r results}
res <- results(dds, contrast = c("group", "DIPG_DMG", "normal pons"))

summary(res)
```

Load gtf file to obtain gene symbols from ensembl gene IDs

```{r get gene symbols}
gtf <- rtracklayer::import(file.path(data_dir,
                                     "gencode.v39.primary_assembly.annotation.gtf.gz"))

gtf_df <- as.data.frame(gtf) %>%
  dplyr::mutate(gene_id = unlist(lapply(strsplit(gene_id, "[.]"), function(x) x[[1]])))

# add gene names to res
res <- res %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  left_join(gtf_df %>% dplyr::select(gene_id,
                                     gene_name),
            by = "gene_id") %>%
  distinct()
```

Generate volcano plot and save

```{r volcanoplot}
volc_plot <- EnhancedVolcano(res,
                             lab = res$gene_name, 
                             x = "log2FoldChange",
                             y = "pvalue",
                             title = "DIPG/DMG versus normal pons",
                             subtitle = NULL,
                             caption = NULL,
                             pCutoff = max(res$pvalue[res$padj < 0.05], na.rm = TRUE),
                             FCcutoff = 2,
                             pointSize = 4,
                             labSize = 3,
                             typeConnectors = "closed",
                             drawConnectors = TRUE,
                             widthConnectors = 1,
                             colConnectors = 'black',
                             max.overlaps = 12)

pdf(file.path(plot_dir,
              "de-mirna-dipg-dmg-versus-normal-volcano-plot.pdf"),
    width = 6, height = 6)

volc_plot

dev.off()
```

Print DIPG/DMG upregulated genes

```{r print up genes}
res %>%
  dplyr::filter(log2FoldChange > 0,
                padj < 0.05) %>%
  arrange(padj)
```

Print DIPG/DMG downregulated genes

```{r print down genes}
res %>%
  dplyr::filter(log2FoldChange < 0,
                padj < 0.05) %>%
  arrange(padj)
```

Plot normalized counts of top 20 DE genes in DIPG/DMG and normal pons

```{r plot cts}
degs <- res %>%
  arrange(pvalue) %>%
  dplyr::filter(padj < 0.05) %>%
  dplyr::select(gene_id, gene_name)

for (i in 1:20){
  
  gene_id <- degs$gene_id[i]
  gene_name <- degs$gene_name[i]
  
  counts_df <- plotCounts(dds,
                          gene = gene_id,
                          intgroup = "group",
                          main = gene_name,
                          returnData = TRUE)
  
  # Convert it into a ggplot
  ggplot(counts_df, aes(x = group, y = count)) +
    geom_jitter(aes(fill = group), width = 0.2, 
                shape = 21,
                size = 3, alpha = 0.85,
                 show.legend = FALSE) +  # Points for individual samples
    geom_boxplot(outlier.shape = NA, alpha = 0.05,
             show.legend = FALSE) +  # Boxplot to show distribution
    scale_y_log10() +  # Log-transform if necessary
    theme_Publication() +
    labs(title = gene_name,
         x = "Group",
         y = "Normalized Count")
  
  ggsave(file.path(plot_dir,
                glue::glue("{gene_name}_cts_dipg_dmg_versus_normal.pdf")),
      width = 5, height = 5)
  
}
```

Save differential expression results

```{r write output}
write_tsv(res, 
          file.path(results_dir,
                    "mrna-differential-expression-deseq2-dipg-dmg-versus-normal.tsv"))
```

Write upregulated and downregulated genes to text files for GSEA:

```{r save gene lists}
upregulated_genes <- res %>%
  dplyr::filter(padj < 0.05,
                log2FoldChange > 0) %>% 
  pull(gene_name)

write_tsv(as.data.frame(upregulated_genes),
          file.path(results_dir, "dmg-upregulated-genes.tsv"))

downregulated_genes <- res %>%
  dplyr::filter(padj < 0.1,
                log2FoldChange < 0) %>% 
  pull(gene_name)

write_tsv(as.data.frame(downregulated_genes),
          file.path(results_dir, "dmg-downregulated-genes.tsv"))
```

Print session info

```{r session info}
sessionInfo()
```
