---
title: 'GO enrichment analysis of DEGs in DIPG/DMG'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2025"
---

This script performs gene ontology (GO) enrichment analysis on DE genes in DIPG/DMG versus normal pons

Load packages

```{r load packages}
library(tidyverse)
library(ggplot2)
library(topGO)
library(org.Hs.eg.db)  
library(AnnotationDbi)
```

Set directory paths

```{r dir paths}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-rna-differential-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

source(file.path(root_dir, "figures", "theme.R"))
```

Set file paths 

```{r file paths}
de_rna_file <- file.path(root_dir, "analyses",
                           "human-rna-differential-expression",
                           "results",
                           "mrna-differential-expression-deseq2-dipg-dmg-versus-normal.tsv")
```

Wrangle data

```{r wrangle data}
de_mrna_df <- read_tsv(de_rna_file)
```

Pull upregulated and downregulated genes

```{r get DEGs}
upregulated_genes <- de_mrna_df %>%
  dplyr::filter(padj < 0.1,
                log2FoldChange > 0) %>%
  pull(gene_name)

downregulated_genes <- de_mrna_df %>%
  dplyr::filter(padj < 0.1,
                log2FoldChange < 0) %>%
  pull(gene_name)
```

Run GO enrichment for upregulated genes

```{r up GO}
all_genes <- de_mrna_df$gene_name[!is.na(de_mrna_df$gene_name)]

is_upregulated <- ifelse(all_genes %in% upregulated_genes,
                         1, 0)
names(is_upregulated) <- all_genes

# Create topGO object, using biological process (BP) ontology terms
GO_data_up <- new("topGOdata", ontology = "BP", allGenes = is_upregulated, 
                                                 geneSel = function(x) x == 1, 
                                                 annot = annFUN.org, mapping = "org.Hs.eg.db", ID = "symbol")

# Run Fisher test
GO_results_up <- runTest(GO_data_up, algorithm = "classic", statistic = "fisher")

# write resutls to go analysis list
GO_table_up <- GenTable(GO_data_up, classicFisher = GO_results_up, topNodes = length(usedGO(GO_data_up)),
                                                     numChar = 100000)

sig_table_up <- GO_table_up %>%
  dplyr::mutate(classicFisher = as.numeric(classicFisher)) %>%
  dplyr::mutate(fdr = p.adjust(classicFisher, method = "BH"),
                 gene_ratio = Significant/Annotated) %>%
  dplyr::arrange(desc(gene_ratio)) %>%
  dplyr::filter(fdr < 0.05)
  
```

Plot most significantly enriched GO terms

```{r sig upregulated terms}
sig_table_up %>%
  dplyr::filter(Significant > 2) %>%
  head(40) %>%
  arrange(gene_ratio) %>%
  dplyr::mutate(Term = fct_relevel(Term, 
                                      unique(Term))) %>%

    ggplot(aes(x = gene_ratio, y = Term,
               size = Significant, colour = -log10(classicFisher))) +
    geom_point(alpha = 0.9) + 
    scale_color_gradient(low = "lightpink", high = "darkred") +
    labs(y = NULL,
         x = "Gene Ratio",
         size = "Count",
         colour = "-log10(p)",
         title = "DIPG/DMG-upregulated genes") +
    theme_Publication()
  

ggsave(file.path(plot_dir, "dipg-dmg-upregulated-genes-go-term-enrichment-dotplot.pdf"),
     width = 10, height = 9)
```
Add list of genes associated with each sig GO term to df

```{r get genes}
genes_per_term <- genesInTerm(GO_data_up)

genes_per_term_df <- enframe(genes_per_term, name = "GO.ID", value = "Genes") %>%
  mutate(Genes = sapply(Genes, function(x) paste(x, collapse = ", ")))

sig_table_up <- sig_table_up %>%
  left_join(genes_per_term_df)
```

Repeat for downregulated genes

```{r down GO}
is_downregulated <- ifelse(all_genes %in% downregulated_genes,
                         1, 0)
names(is_downregulated) <- all_genes

# Create topGO object, using biological process (BP) ontology terms
GO_data_down <- new("topGOdata", ontology = "BP", allGenes = is_downregulated, 
                                                 geneSel = function(x) x == 1, 
                                                 annot = annFUN.org, mapping = "org.Hs.eg.db", ID = "symbol")

# Run Fisher test
GO_results_down <- runTest(GO_data_down, algorithm = "classic", statistic = "fisher")

# write resutls to go analysis list
GO_table_down <- GenTable(GO_data_down, classicFisher = GO_results_down, topNodes = length(usedGO(GO_data_down)),
                                                     numChar = 100000)

sig_table_down <- GO_table_down %>%
  dplyr::mutate(classicFisher = as.numeric(classicFisher)) %>%
  dplyr::mutate(fdr = p.adjust(classicFisher, method = "BH"),
                 gene_ratio = Significant/Annotated) %>%
  dplyr::arrange(desc(gene_ratio)) %>%
  dplyr::filter(fdr < 0.05)
```

There are currently no significantly enriched pathways among DIPG/DMG downregulated genes

Save upregulated gene GO term enrichment results:

```{r}
write_tsv(sig_table_up,
          file.path(results_dir,
                    "enriched-go-terms-dipg-dmg-upregulated-genes.tsv"))
```