---
title: 'GSVA '
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2025"
---

This script performs gene set variation analysis (GSVA) on DIPG/DMG and normal pons expression data, and performs differential pathway expression

Load packages

```{r load packages}
library(tidyverse)
library(readr)
library(tibble)
library(msigdbr) 
library(GSVA) 
```

Set directory paths

```{r dir paths}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-rna-differential-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")
```

Set file paths

```{r file paths}
expr_file <- file.path(data_dir,
                       "30-992989426-RNA-TPM_values.csv")
```

Load expression data and kegg pathways

```{r wrangle data}
expr <- read_csv(expr_file) %>%
  column_to_rownames("...1")

human_kegg  <- msigdbr::msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG") 

```

Create a list of KEGG pathways, each of which is a list of genes

```{r get pathway genes}
human_kegg_twocols <- human_kegg %>% dplyr::select(gs_name, ensembl_gene)
human_kegg_list    <- base::split(human_kegg_twocols$ensembl_gene, list(human_kegg_twocols$gs_name))
```

Run GSVA

```{r gsva}
expr_log2_matrix <- as.matrix(log2(expr + 1))

gsva_scores_param <- gsvaParam(expr_log2_matrix,
                                 geneSets = human_kegg_list,
                                 kcdf = "Gaussian",
                                 assay = NA_character_,
                                 annotation = NA_character_,
                                 tau = 1,
                                 minSize = 1, 
                                 maxSize = 1500, ## Arguments from K. Rathi
                                 maxDiff = TRUE)

gsva_scores <- gsva(gsva_scores_param, verbose = TRUE)

```

Run differential expression analysis on pathway GSVA scores

```{r differential expression}
group <- ifelse(grepl("CC1", colnames(gsva_scores)), "2_DIPG/DMG", "1_pons")

mod <- model.matrix(~ factor(group))
colnames(mod) <- c("intercept", "2_DIPG/DMG")
rownames(mod) <- colnames(gsva_scores)
fit <- limma::lmFit(object = gsva_scores, design = mod)
fit <- limma::eBayes(fit)
toptable_output <- limma::topTable(fit, coef = 2, n = Inf)

toptable_output <- toptable_output %>%
  rownames_to_column("Geneset") %>%
  arrange(P.Value)
```

Pull DE pathways at nominal p < 0.05 and create expression matrix for plotting

```{r prepare plot matrix}
DEpwys <- toptable_output %>%
      filter(P.Value < 0.05) %>%
      arrange(P.Value) %>%
      pull(Geneset) 

DEpwys_es <- gsva_scores[DEpwys, ]
rownames(DEpwys_es) <- gsub("KEGG_|HALLMARK_", "", rownames(DEpwys_es))
rownames(DEpwys_es) <- gsub("_", " ", rownames(DEpwys_es))

# update sample names for plotting
colnames(DEpwys_es) <- c("DMG-1", "DMG-2", "normal pons")
```

Plot expression heatmap of signifiantly DE pathways

```{r plot}
color_palette <- colorRampPalette(c("blue", "white", "darkorange"))
heat_colors <- color_palette(6)

pheatmap::pheatmap(DEpwys_es, 
                   fontsize = 8,
                   treeheight_row = 10, 
                   show_colnames = TRUE,
                   cluster_cols = FALSE, 
                   color = heat_colors,
                   name = "GSVA score",
                   filename = file.path(plot_dir, "de-pathway-expr-heatmap.pdf"), 
                   width = 6, height = 6)
```

Save results files

```{r save outputs}
gsva_scores %>%
  as.data.frame() %>%
  rownames_to_column("KEGG_pathway") %>%
  write_tsv(file.path(results_dir, "kegg-gsva-scores.tsv"))

write_tsv(toptable_output,
          file.path(results_dir,
                    "kegg-pathway-differential-expression-dipg-dmg-versus-normal-pons.tsv"))
```

Print session info

```{r session info}
sessionInfo()
```