---
title: "Differential RNA Expression Analysis"
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Bicna Song
date: "`r Sys.Date()`"
---

This script performs differential expression analysis of RNA-seq data using DESeq2.

The goal is to identify differentially expressed genes for each brain tumor histology by comparing:

- Tumor vs. Tumor-Adjacent Normal samples (e.g., DIPG vs DIPG-adjacent normal)
- Tumor vs. Disease-Free Normal brain samples (B1–B5)

### Setup

```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(DESeq2)
library(tidyverse)
library(rprojroot)
library(readr)
library(EnhancedVolcano)
library(ashr)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-rna-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# File path for counts
gene_cts_file <- file.path(data_dir, "gene-counts-rsem-expected_count-collapsed.all.rds")

# File path for histology
histology_file <- file.path(root_dir, "analyses", "histology-preprocessing", "results", "histologies.tsv")
```

### Data Wrangling

```{r load-and-clean-histology}
# define which histologies to keep
keep_histologies <- c("DIPG or DMG", "EPN", "MB", "LGG", "Normal")
batch1_ids <- c("2-1578-CC1", "3-1890-CC1")

histology_df <- read_delim(histology_file) %>%
  filter(experimental_strategy == "RNA-Seq") %>%
  mutate(histology = coalesce(histology, "Normal"),
         sample_id = case_when(
           external_sample_id == "1516-T" ~ "7316-NA",
           external_sample_id %in% paste0("B", 1:5) ~ external_sample_id,
           TRUE ~ sample_id),
         batch = if_else(external_sample_id %in% batch1_ids, "batch1", "batch2"),
         condition = case_when(
           sample_type %in% c("Tumor", "Normal") ~ histology,
           sample_type == "Normal-Tumor Adjacent"~ paste0(histology, "-adjacent Normal"),
           TRUE ~ NA_character_
           ),
         histology_condition = case_when(
           sample_type %in% c("Tumor", "Normal") ~ histology,
           sample_type == "Normal-Tumor Adjacent" ~ "Normal-Tumor Adjacent",
           TRUE ~ NA_character_
           )
         ) %>%
  filter(histology %in% keep_histologies)
```

```{r load-and-filter-counts}
# load & filter genes with at least two non-zero samples
gene_cts <- readRDS(gene_cts_file) %>%
  filter(rowSums(across(everything(), ~ . != 0)) >= 2)

# subset to samples in histology
valid_ids <- histology_df$Bioassay_ID
gene_cts <- round(gene_cts[, intersect(colnames(gene_cts), valid_ids), drop = FALSE])
```

```{r align-metadata-to-counts}
histology_df <- histology_df %>%
  arrange(match(Bioassay_ID, names(gene_cts)))
```

### Perform Differential Expression for Each Histology

```{r run-de-by-histology, message=FALSE, warning=FALSE}
normal_ids <- c("B1", "B2", "B3", "B4", "B5")
histologies <- c("DIPG or DMG", "EPN", "MB")

for (tumor in histologies) {
  adj_label <- paste0(tumor, "-adjacent Normal")

  # Subset metadata
  meta_adj <- histology_df %>% filter(condition %in% c(tumor, adj_label))
  meta_healthy <- histology_df %>% filter(external_sample_id %in% c(normal_ids, external_sample_id[condition == tumor])) %>% 
    filter(condition %in% c(tumor, "Normal"))

  # Select count data
  counts_adj <- gene_cts[, meta_adj$Bioassay_ID]
  counts_healthy <- gene_cts[, meta_healthy$Bioassay_ID]

  # Tumor vs Adjacent Normal
  design_adj <- if (tumor == "DIPG or DMG") ~ condition else ~ condition
  dds_adj <- DESeqDataSetFromMatrix(countData = counts_adj, colData = meta_adj, design = design_adj)
  dds_adj$condition <- droplevels(dds_adj$condition)
  dds_adj <- DESeq(dds_adj)
  res_adj <- results(dds_adj, contrast = c("condition", tumor, adj_label))
  res_adj <- lfcShrink(dds_adj, contrast = c("condition", tumor, adj_label), res = res_adj, type = "ashr")
  write_csv(as.data.frame(res_adj) %>% rownames_to_column("gene_symbol"),
          file.path(results_dir, paste0("DESeq2_", tumor, "_vs_adjNormal.csv")))

  # Tumor vs Healthy Normal
  design_healthy <- if (tumor == "DIPG or DMG") ~ condition else ~ condition
  dds_healthy <- DESeqDataSetFromMatrix(countData = counts_healthy, colData = meta_healthy, design = design_healthy)
  dds_healthy$condition <- droplevels(dds_healthy$condition)
  dds_healthy <- DESeq(dds_healthy)
  res_healthy <- results(dds_healthy, contrast = c("condition", tumor, "Normal"))
  res_healthy <- lfcShrink(dds_healthy, contrast = c("condition", tumor, "Normal"), res = res_healthy, type = "ashr")
  write_csv(as.data.frame(res_healthy) %>% rownames_to_column("gene_symbol"),
          file.path(results_dir, paste0("DESeq2_", tumor, "_vs_healthyNormal.csv")))
}
```

Handle LGG separately: only Tumor vs Healthy Normal

```{r run-de-lgg, message=FALSE, warning=FALSE}
# Handle LGG separately: only Tumor vs Healthy Normal
tumor <- "LGG"
meta_lgg <- histology_df %>%
  filter(external_sample_id %in% c(normal_ids, external_sample_id[condition == tumor])) %>%
  filter(condition %in% c(tumor, "Normal"))
counts_lgg <- gene_cts[, meta_lgg$Bioassay_ID]

design_lgg <- ~ condition
dds_lgg <- DESeqDataSetFromMatrix(countData = counts_lgg, colData = meta_lgg, design = design_lgg)

dds_lgg$condition <- droplevels(dds_lgg$condition)
dds_lgg <- DESeq(dds_lgg)

res_lgg <- results(dds_lgg, contrast = c("condition", tumor, "Normal"))
res_lgg <- lfcShrink(dds_lgg, contrast = c("condition", tumor, "Normal"), res = res_lgg, type = "ashr")

write_csv(as.data.frame(res_lgg) %>% rownames_to_column("gene_symbol"),
          file.path(results_dir, paste0("DESeq2_", tumor, "_vs_healthyNormal.csv")))
```

### Save Significant DE Results Separately

```{r save-sig-de-res}
all_histologies <- c("DIPG or DMG", "EPN", "MB", "LGG")

for (tumor in all_histologies) {
  comparison_types <- if (tumor == "LGG") "healthyNormal" else c("adjNormal", "healthyNormal")
  
  for (type in comparison_types) {
    file_path <- file.path(results_dir, paste0("DESeq2_", tumor, "_vs_", type, ".csv"))
    if (file.exists(file_path)) {
      res_df <- read_csv(file_path)
      sig_df <- res_df %>% filter(!is.na(padj) & padj < 0.05)
      write_csv(sig_df, file.path(results_dir, paste0("DESeq2_", tumor, "_vs_", type, "_sig.csv")))
    }
  }
}
```

### Summary Table: Number of Up/Down-regulated genes

```{r summary-table}
summary_df <- tibble()

for (tumor in all_histologies) {
  comparison_types <- if (tumor == "LGG") "healthyNormal" else c("adjNormal", "healthyNormal")
  
  for (type in comparison_types) {
    file_path <- file.path(results_dir, paste0("DESeq2_", tumor, "_vs_", type, ".csv"))
    if (file.exists(file_path)) {
      res_df <- read_csv(file_path) %>%
        filter(!is.na(padj) & padj < 0.05)
      up <- sum(res_df$log2FoldChange > 0)
      down <- sum(res_df$log2FoldChange < 0)
      summary_df <- bind_rows(summary_df, 
        tibble(Comparison = paste0(tumor, " vs ", type), 
               Upregulated = up, 
               Downregulated = down))
    }
  }
}

summary_df

# Save to CSV
write_csv(summary_df, file.path(results_dir, "DE_summary_counts.csv"))
```

### Volcano Plots for DE Comparisons

```{r volcano-plots, message=FALSE, warning=FALSE}
for (tumor in all_histologies) {
  comparison_types <- if (tumor == "LGG")
    "healthyNormal"
  else
    c("adjNormal", "healthyNormal")
  
  for (type in comparison_types) {
    file_path <- file.path(results_dir, paste0("DESeq2_", tumor, "_vs_", type, ".csv"))
    if (file.exists(file_path)) {
      res_df <- read_csv(file_path)
      rownames(res_df) <- res_df$gene_symbol %||% rownames(res_df)
      p_cut <- max(res_df$pvalue[res_df$padj < 0.05], na.rm = TRUE)
      comp_label <- paste(tumor, "vs", gsub("Normal", "normal", type))
      
      top_labels <- res_df %>%
        filter(!is.na(padj)) %>%
        arrange(padj, desc(abs(log2FoldChange))) %>%
        slice_head(n = 10) %>%
        pull(gene_symbol)
      
      plot <- EnhancedVolcano(
        res_df,
        lab = rownames(res_df),
        selectLab = top_labels,
        x = "log2FoldChange",
        y = "pvalue",
        title = comp_label,
        subtitle = NULL,
        caption = NULL,
        pCutoff = p_cut,
        FCcutoff = 2,
        pointSize = 3.5,
        labSize = 4.5,
        typeConnectors = "closed",
        drawConnectors = TRUE,
        widthConnectors = 0.75,
        colConnectors = 'grey30',
        max.overlaps = 10
      )
      
      ggsave(
        file.path(plot_dir, paste0(
          "volcano_", tumor, "_vs_", type, ".pdf"
        )),
        plot = plot,
        width = 8,
        height = 6
      )
      
    }
    
  }
  
}
```

### Summarize: Count number of Tumor and Adjacent Normal per sample_id and histology

```{r pair-status}
paired_info <- histology_df %>%
  group_by(histology, sample_id) %>%
  summarize(
    tumor = any(sample_type == "Tumor"),
    adjacent_normal = any(sample_type == "Normal-Tumor Adjacent"),
    .groups = "drop"
  ) %>%
  mutate(pair_status = case_when(
    tumor & adjacent_normal ~ "Paired",
    TRUE ~ "Unpaired"
  ))

# Create table of paired/unpaired counts per histology
summary_table <- paired_info %>%
  group_by(histology, pair_status) %>%
  summarize(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = pair_status, values_from = n, values_fill = 0)

print(summary_table)
```

### Paired differential expression analysis

```{r run-paired-de-by-histology}
for (hist in all_histologies) {
  
  paired_patients <- paired_info %>%
    filter(histology == hist, pair_status == "Paired") %>%
    pull(sample_id)
  
  if (length(paired_patients) < 2) {
    message(paste("Skipping", hist, "due to insufficient paired patients"))
    next
  }
  
  paired_samples_metadata <- histology_df %>%
    filter(histology == hist, sample_id %in% paired_patients) %>%
    filter(sample_type %in% c("Tumor", "Normal-Tumor Adjacent"))
  
  paired_external_sample_ids <- paired_samples_metadata$Bioassay_ID
  
  gene_cts_paired <- gene_cts[, paired_external_sample_ids]
  
  # Confirm external_sample_id matches column names
  stopifnot(identical(paired_samples_metadata$Bioassay_ID, colnames(gene_cts_paired)))
  
  # Create DESeq2 object
  dds <- DESeqDataSetFromMatrix(
    countData = gene_cts_paired,
    colData = paired_samples_metadata,
    design = ~ sample_id + sample_type
  )
  
  # run DESeq
  dds <- DESeq(dds)
  
  # Run results: Tumor vs Normal-Tumor Adjacent
  res <- results(dds, contrast = c("sample_type", "Tumor", "Normal-Tumor Adjacent"))
  res_adj <- lfcShrink(dds, contrast = c("sample_type", "Tumor", "Normal-Tumor Adjacent"), res = res, type = "ashr")
  
  # Save full and significant results
  res_df <- as.data.frame(res_adj) %>% 
    rownames_to_column("gene_symbol")
  
  write_csv(res_df, file.path(results_dir, paste0("DESeq2_", hist, "_paired_full.csv")))
  
  res_sig <- res_df %>%
    filter(!is.na(padj) & padj < 0.05)
  
  write_csv(res_sig, file.path(results_dir, paste0("DESeq2_", hist, "_paired_sig.csv")))
  
  message(paste("Completed DE analysis for:", hist))
}

```

### Summary Table: Number of Up/Down-regulated miRNAs for paired analysis

```{r}
summary_paired_df <- tibble()

for (hist in all_histologies) {
  file_path <- file.path(results_dir, paste0("DESeq2_", hist, "_paired_full.csv"))
  
  if (file.exists(file_path)) {
      res_df <- read_csv(file_path) %>%
        filter(!is.na(padj) & padj < 0.05)
      up <- sum(res_df$log2FoldChange > 0)
      down <- sum(res_df$log2FoldChange < 0)
      summary_paired_df <- bind_rows(summary_paired_df, 
        tibble(Comparison = paste0(hist, " vs ", "adjNormal"), 
               Upregulated = up, 
               Downregulated = down))
  } else {
    message(paste("Result file for", hist, "not found."))
  }
}

# View the summary table
summary_paired_df

# Save to CSV
write_csv(summary_paired_df, file.path(results_dir, "DE_summary_counts_paired.csv"))
```

### Volcano Plots for paired DE Comparisons

```{r}
for (tumor in all_histologies) {
  file_path <- file.path(results_dir, paste0("DESeq2_", tumor, "_paired_full.csv"))
  
  if (file.exists(file_path)) {
    res_df <- read_csv(file_path)
    rownames(res_df) <- res_df$gene_symbol %||% rownames(res_df)
    p_cut <- max(res_df$pvalue[res_df$padj < 0.05], na.rm = TRUE)
    comp_label <- paste(tumor, "Paired Analysis")
    top_labels <- res_df %>%
      filter(!is.na(padj)) %>%
      arrange(padj, desc(abs(log2FoldChange))) %>%
      slice_head(n = 10) %>%
      pull(gene_symbol)
    
    plot <- EnhancedVolcano(
      res_df,
      lab = rownames(res_df),
      selectLab = top_labels,
      x = "log2FoldChange",
      y = "pvalue",
      title = comp_label,
      subtitle = NULL,
      caption = NULL,
      pCutoff = p_cut,
      FCcutoff = 2,
      pointSize = 3.5,
      labSize = 4.5,
      typeConnectors = "closed",
      drawConnectors = TRUE,
      widthConnectors = 0.75,
      colConnectors = 'grey30',
      max.overlaps = 10
    )
    
    ggsave(
      file.path(plot_dir, paste0("volcano_", tumor, "_paired.pdf")),
      plot = plot,
      width = 8,
      height = 6
    )
  }
}
```

### Session Info

```{r session-info}
sessionInfo()
```
