---
title: "GSVA of RNA-Seq TPM Using GO–Biological Process"
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Bicna Song
date: "`r Sys.Date()`"
---

### Overview

Perform Gene Set Variation Analysis (GSVA) on RSEM-derived TPM expression for multiple brain tumor histologies, using GO Biological Process terms, and assess differential pathway activity.

### Setup

```{r setup, message=FALSE, warning=FALSE}
set.seed(2025)

# Core libraries
library(GSVA)
library(msigdbr)
library(clusterProfiler)
library(limma)
library(dplyr)
library(readr)
library(tibble)
library(ggplot2)
library(pheatmap)
library(rprojroot)
library(tidyverse)

# Define colors
heat_colors <- colorRampPalette(c("blue","white","darkorange"))(6)
```

### Set directory paths

```{r dir-paths}
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-rna-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")
```

### Load file

```{r load-file}
# RDS with TPM expression
tpm_rds <- file.path(data_dir, "gene-expression-rsem-tpm-collapsed.all.rds")

# Load a matrix: genes × samples
expr_tpm <- readRDS(tpm_rds)

# Remove 5-1234-left-pons showing significant tumor infiltration
expr_tpm <- expr_tpm[, !(colnames(expr_tpm) == "BA_GBVGBCJG")]

# Log2‐transform
expr_mat <- as.matrix(log2(expr_tpm + 1))
```

### Load GO–BP Gene Sets

```{r}
go_bp_df <- msigdbr(
  species    = "Homo sapiens",
  category   = "C5",
  subcategory= "BP"
) %>%
  select(gs_name, gene_symbol)

go_bp_list <- split(go_bp_df$gene_symbol, go_bp_df$gs_name)
```

### Run GSVA

```{r run-gsva, warning=FALSE}
gsva_scores_param <- gsvaParam(expr_mat,
                               geneSets = go_bp_list,
                               kcdf = "Gaussian",
                               maxSize = 1500)

gsva_scores <- gsva(gsva_scores_param, verbose = TRUE)
```
```{r}
gsva_scores %>%
  as.data.frame() %>%
  rownames_to_column("Pathway") %>%
  write_tsv(file.path(results_dir, "gobp-gsva-scores.tsv"))
```

### Run differential expression analysis on pathway GSVA scores

```{r}
all_hist <- c("ATRT","DIPG or DMG","EPN","HGG","LGG","MB")
normal_ids <- c("B1","B2","B3","B4","B5")
```

```{r load-histology}
# Load and annotate histology file
histology_df <- read_delim(file.path(root_dir, "analyses", "histology-preprocessing", "results", "histologies.tsv")) %>%
  filter(experimental_strategy == "RNA-Seq" & external_sample_id != "5-1234-left-pons") %>%
  # Fill in any missing histology with "Normal"
  mutate(histology = if_else(is.na(histology), "Normal", histology)) %>%
  mutate(condition = case_when(
      sample_type %in% c("Tumor", "Normal") ~ histology,
      sample_type == "Normal-Tumor Adjacent" ~ paste0(histology, "-adjacent Normal"),
      TRUE ~ NA_character_
    ),
    histology_condition = case_when(
      sample_type %in% c("Tumor", "Normal") ~ histology,
      sample_type == "Normal-Tumor Adjacent" ~ "Normal-Tumor Adjacent",
      TRUE ~ NA_character_
    )
  )
```

```{r pair-status}
paired_info <- histology_df %>%
  group_by(histology, sample_id) %>%
  summarize(
    tumor = any(sample_type == "Tumor"),
    adjacent_normal = any(sample_type == "Normal-Tumor Adjacent"),
    .groups = "drop"
  ) %>%
  mutate(pair_status = case_when(
    tumor & adjacent_normal ~ "Paired",
    TRUE ~ "Unpaired"
  ))

# Create table of paired/unpaired counts per histology
summary_table <- paired_info %>%
  group_by(histology, pair_status) %>%
  summarize(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = pair_status, values_from = n, values_fill = 0)

print(summary_table)
```

Identify keywords in GO terms for plotting (provided by Dalia)

```{r keywords}
keywords <- "immunity|T cell|immune|leukocyte|lymphocyte|myeloid|signal transduction|transforming growth factor beta|MAPK cascade|autophagy|cell death|glial cell|Ras|small GTPase|secretion|intracellular signal transduction|phosphorylation|cell comm|cell adhesion|defense"
```

```{r de-gsva, message=TRUE, warning=FALSE}
for (hist in all_hist) {
  message("Histology: ", hist)
  # Determine paired vs. unpaired samples using paired_info
  paired_sample_ids <- paired_info %>%
    filter(histology == hist, pair_status == "Paired") %>%
    pull(sample_id)
  # Map to Bioassay_IDs
  paired_ids <- histology_df %>%
    filter(sample_id %in% paired_sample_ids) %>%
    pull(Bioassay_ID)

  # Unpaired (tumor vs healthy normals)
  healthy_ids <- histology_df %>%
    filter(histology==hist & sample_type=="Tumor" | external_sample_id %in% normal_ids) %>% pull(Bioassay_ID)
  
  for (contrast in c("paired","healthyNormal")) {
    ids <- if (contrast=="paired") paired_ids else healthy_ids
    if (length(ids) < 2) next
    mat <- gsva_scores[, ids, drop=FALSE]
    meta <- histology_df %>%
      filter(Bioassay_ID %in% ids) %>%
      mutate(condition = if_else(sample_type == "Tumor", hist, "Normal")) %>%
      # ensure 'condition' is a factor with 'Normal' baseline
      mutate(condition = factor(condition, levels = c("Normal", hist)))

    # skip if any condition has fewer than two samples
    cond_counts <- table(meta$condition)
    if (any(cond_counts < 2)) {
      message("   ✗ Insufficient replicates for ", hist, "/", contrast, ": ", paste(names(cond_counts), cond_counts, sep="=", collapse=", "))
      next
    }
    design <- model.matrix(~ condition, data=meta)
    fit <- lmFit(mat, design) %>% eBayes()
    res <- topTable(fit, coef = 2, number = Inf) %>%
      rownames_to_column("Pathway")
    
    if (nrow(res) > 0) {
      out_tsv <- paste0(gsub("[ ]", "_", hist), "_" , contrast, "_GSVA_DE_results.tsv")
      write_tsv(res, file.path(results_dir, out_tsv))
      message("   ✓ Saved results: ", out_tsv)

      # filter significant pathways by keyword
      sig <- res %>%
        filter(P.Value < 0.05) %>%
        filter(stringr::str_detect(Pathway, regex(keywords, ignore_case = TRUE))) %>%
        arrange(P.Value) %>%
        slice_head(n = 30) %>%
        pull(Pathway)
      
      if (length(sig) > 1) {
        # reorder columns to show Tumor first, then Normal
        col_order <- meta %>% arrange(desc(sample_type)) %>% pull(Bioassay_ID)
        hm <- mat[sig, col_order, drop = FALSE]
        rownames(hm) <- gsub("^GOBP_", "", rownames(hm))
        colnames(hm) <- meta %>% arrange(desc(sample_type)) %>% pull(external_sample_id)
        pheatmap(hm,
                 fontsize = 8,
                 treeheight_row = 10,
                 color = colorRampPalette(c("blue", "white", "darkorange"))(6),
                 cluster_cols = FALSE,
                 filename = file.path(plot_dir,
                                      paste0("heatmap_", gsub("[ ]", "_", hist), "_", contrast, ".pdf")),
                 width = 12)
        message("   ✓ Plotted heatmap for ", hist, "/", contrast)
      }
    }
  }
}
```


### Session Info

```{r session-info}
sessionInfo()
```



