---
title: "Correlation between DE miRNA expression and GSVA score"
output:
  html_document:
    toc: true
    toc_float: true
author: "Bicna Song"
date: "`r Sys.Date()`"
---

### Overview

Correlate DE miRNA expression with GSVA pathway scores in DIPG/DMG tumors; generate full and significant-pairs heatmaps with stable annotations.


### Setup

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(tibble)
library(ggplot2)
library(pheatmap)
library(rprojroot)
library(stringr)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(cluster)
```

### Paths and Files

```{r paths}
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-rna-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")
input_dir <- file.path(analysis_dir, "input")

miRNA_tpm <- file.path(input_dir, "mirna-tpm.rds")
histology_file <- file.path(root_dir, "analyses", "histology-preprocessing", "results", "histologies.tsv")

all_hist <- c("DIPG or DMG","EPN","LGG","MB")
normal_ids <- c("B1","B2","B3","B4","B5")
```

### Load Data

```{r load-data}
# miRNA TPM
mirna_tpm <- readRDS(miRNA_tpm)
mirna_tpm <- log2(as.matrix(mirna_tpm) + 1)

# Histology
histology_df <- read_delim(histology_file, show_col_types = FALSE)
```

### Load precomputed GSVA scores

```{r load-gsva-scores}
# read in GSVA scores from previous run
gsva_scores <- read_tsv(file.path(results_dir, "gobp-gsva-scores.tsv"), col_types = cols()) %>%
  column_to_rownames("Pathway") %>%
  as.matrix()
```


### Select DIPG/DMG tumor samples and subset matrices

```{r select-samples}
dipg_tumor_bioassay_ids <- histology_df %>%
  filter(experimental_strategy == "RNA-Seq",
         sample_type == "Tumor",
         histology == "DIPG or DMG") %>%
  pull(Bioassay_ID)

dipg_tumor_ext_ids <- histology_df %>%
  filter(experimental_strategy == "RNA-Seq",
         sample_type == "Tumor",
         histology == "DIPG or DMG") %>%
  pull(external_sample_id)

# Subset GSVA and miRNA matrices
pathway_scores <- gsva_scores[, dipg_tumor_bioassay_ids, drop=FALSE]
miRNA_expr <- mirna_tpm  [, dipg_tumor_ext_ids, drop=FALSE]
```

### Correlate miRNA Expression with Pathway Scores

```{r correlation, message=FALSE, warning=FALSE}
# Select DE miRNAs of interest (170 genes)
DIPG_sig_DE_miRNA_list <- read.csv(file = "input/DIPG or DMG_sig_DE_miRNA_list.csv")

de_mirnas <- DIPG_sig_DE_miRNA_list$miRNA
miRNA_expr_sub <- miRNA_expr[rownames(miRNA_expr) %in% de_mirnas, , drop=FALSE]

# Spearman correlation for all pairs
rho_mat <- cor(t(miRNA_expr_sub), t(pathway_scores), method = "spearman")
rho_mat_filtered <- rho_mat[rowSums(is.na(rho_mat)) != ncol(rho_mat), ]

# p-values (approx)
n <- ncol(miRNA_expr_sub)
t_val <- rho_mat_filtered * sqrt((n-2)/(1 - rho_mat_filtered^2))
pval_mat <- 2 * pt(-abs(t_val), df = n-2)

# Long format with BH adjustment per miRNA
corr_df <- as.data.frame(as.table(rho_mat_filtered)) %>%
  rename(miRNA = Var1, Pathway = Var2, rho = Freq) %>%
  mutate(pval = as.vector(pval_mat)) %>%
  group_by(miRNA) %>%
  mutate(padj = p.adjust(pval, method = "BH")) %>%
  ungroup() %>%
  arrange(miRNA, desc(abs(rho)))
```

### Heatmap: all pairs (filtered to non-NA rows)

```{r}
rho_mat_filtered <- corr_df %>%
  select(miRNA, Pathway, rho) %>%
  pivot_wider(names_from = miRNA, values_from = rho) %>%
  column_to_rownames("Pathway") %>%
  as.matrix()

# Annotation mapping for "all pairs" view 
sample_anno <- data.frame(miRNA = colnames(rho_mat_filtered)) %>%
  left_join(DIPG_sig_DE_miRNA_list, by = "miRNA")

sample_anno$Category  <- factor(sample_anno$Category,
                                levels = c("Paired T vs AN", "Shared", "T vs HN"))
sample_anno$direction <- factor(sample_anno$direction,
                                levels = c("up", "down"))

comp_cols <- c("Paired T vs AN" = "#1b9e77",
               "Shared"         = "#d95f02",
               "T vs HN"        = "#7570b3")
dir_cols  <- c(up = "#E41A1C", down = "#377EB8")

# Color mapping for correlation
col_fun <- colorRamp2(c(-1, 0, 1), c("blue", "white", "red"))

ha_all <- HeatmapAnnotation(
  Comparison = sample_anno$Category,
  Direction  = sample_anno$direction,
  col = list(Comparison = comp_cols, Direction = dir_cols),
  annotation_name_gp = gpar(fontsize = 10),
  annotation_legend_param = list(
    Comparison = list(title = "Comparison", at = levels(sample_anno$Category)),
    Direction  = list(title = "Direction",  at = levels(sample_anno$direction))
  )
)

ht1 <- Heatmap(rho_mat_filtered,
               cluster_columns = TRUE, 
               clustering_distance_columns = "euclidean",
               clustering_method_columns = "complete",
               cluster_rows = TRUE,    
               clustering_distance_rows = "euclidean",
               clustering_method_rows = "complete",
               col = col_fun, 
               na_col = "gray", 
               name = "cor",
               show_row_names = FALSE, 
               show_column_names = FALSE,
               top_annotation = ha_all,
               use_raster = FALSE,
               heatmap_legend_param = list(legend_gp = gpar(fontsize = 10),
                                           labels_gp = gpar(fontsize = 10)))
print(ht1)

```

```{r}
pdf(file.path(plot_dir, "heatmap_all.pdf"))
print(ht1)
dev.off()
```

# significant miRNA–pathway pairs

```{r}
signif_corr <- corr_df %>%
  filter(abs(rho) > 0.5, padj < 0.1)

sig_pathway <- unique(signif_corr$Pathway)
sig_mirna   <- unique(signif_corr$miRNA)

corr_df_sig <- corr_df %>%
  filter(Pathway %in% sig_pathway, miRNA %in% sig_mirna)

rho_mat_sig <- corr_df_sig %>%
  select(miRNA, Pathway, rho) %>%
  pivot_wider(names_from = miRNA, values_from = rho) %>%
  column_to_rownames("Pathway") %>%
  as.matrix()
```

```{r}
sample_anno <- data.frame(miRNA = colnames(rho_mat_sig)) %>%
  left_join(DIPG_sig_DE_miRNA_list, by = "miRNA")

sample_anno$Category  <- factor(sample_anno$Category,
                                levels = c("Paired T vs AN", "Shared", "T vs HN"))
sample_anno$direction <- factor(sample_anno$direction,
                                levels = c("up", "down"))
```

```{r}
# define sample annotation 
ha_sig <- HeatmapAnnotation(
  Comparison = sample_anno$Category,
  Direction  = sample_anno$direction,
  col = list(Comparison = comp_cols, Direction = dir_cols),
  annotation_name_gp = gpar(fontsize = 10),
  annotation_legend_param = list(
    Comparison = list(title = "Comparison", at = levels(sample_anno$Category)),
    Direction  = list(title = "Direction",  at = levels(sample_anno$direction))
  )
)
```

```{r}
# ---- Find optimal k (silhouette) for rows and columns ----
dist_rows <- dist(rho_mat_sig, method = "euclidean")
dist_cols <- dist(t(rho_mat_sig), method = "euclidean")
hc_rows   <- hclust(dist_rows, method = "complete")
hc_cols   <- hclust(dist_cols, method = "complete")

# silhouette-based optimal k (deterministic)
opt_k_sil <- function(dist_mat, hc, k_min = 2, k_max = 10) {
  n <- attr(dist_mat, "Size"); k_max <- min(k_max, n - 1L)
  if (n <= k_min) return(1L)
  sil <- rep(NA_real_, k_max)
  for (k in k_min:k_max) {
    cl <- cutree(hc, k)
    sil[k] <- mean(silhouette(cl, dist_mat)[, 3])
  }
  best <- which.max(sil); if (length(best)==0 || is.na(sil[best])) 1L else best
}
opt_row_k <- opt_k_sil(dist_rows, hc_rows)
opt_col_k <- opt_k_sil(dist_cols, hc_cols)

row_groups <- cutree(hc_rows, k = opt_row_k)
col_groups <- cutree(hc_cols, k = opt_col_k)

message("Optimal clusters — rows: ", opt_row_k, " | columns: ", opt_col_k)
```

```{r}
ht2 <- Heatmap(
  rho_mat_sig,
  cluster_rows    = as.dendrogram(hc_rows),
  cluster_columns = as.dendrogram(hc_cols),
  row_split = opt_row_k,
  column_split = opt_col_k,
  col = col_fun, na_col = "gray", name = "cor",
  show_row_names = FALSE, show_column_names = FALSE,
  top_annotation = ha_sig
)
draw(ht2)
```

```{r}
pdf(file.path(plot_dir, "heatmap_sig_cluster.pdf"))
print(ht2)
dev.off()
```

### Session Info

```{r session-info}
sessionInfo()
```
