---
title: "Correlation between DE miRNA expression and GSVA score"
output:
  html_document:
    toc: true
    toc_float: true
author: "Bicna Song"
date: "`r Sys.Date()`"
params:
  histology: "DIPG or DMG" # MB
  de_list_file: "input/DIPG or DMG_sig_DE_miRNA_list.csv" # input/MB_sig_DE_miRNA_list.csv
  exclude_miRNA: !r character()
  # exclude_miRNA: "hsa-miR-378d" for MB
---

### Overview

Correlate DE miRNA expression with GSVA pathway scores in the selected histology; generate full and significant-pairs heatmaps with stable annotations.


### Setup

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(tibble)
library(ggplot2)
library(pheatmap)
library(rprojroot)
library(stringr)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(cluster)
library(purrr)
library(factoextra) # add to docker
```

### Paths and Files

```{r define-paths}
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-rna-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")
input_dir <- file.path(analysis_dir, "input")

miRNA_tpm <- file.path(input_dir, "mirna-tpm.rds")
histology_file <- file.path(root_dir, "analyses", "histology-preprocessing", "results", "histologies.tsv")

# Precomputed GSVA scores
gsva_file <- file.path(results_dir, "gobp-gsva-scores.tsv")

# Parameters
histology_label <- params$histology
de_list_file    <- params$de_list_file
exclude_miRNA   <- params$exclude_miRNA
```

### Load Data

```{r load-data}
# miRNA TPM
mirna_tpm <- readRDS(miRNA_tpm)
mirna_tpm <- log2(as.matrix(mirna_tpm) + 1)

# Histology
histology_df <- read_delim(histology_file, show_col_types = FALSE)

# GSVA scores
gsva_scores <- read_tsv(gsva_file, show_col_types = FALSE) %>%
  column_to_rownames("Pathway") %>%
  as.matrix()
```

### Select tumor samples and subset matrices

```{r select-samples}
tumor_bioassay_ids <- histology_df %>%
  filter(experimental_strategy == "RNA-Seq",
         sample_type == "Tumor",
         histology == histology_label) %>%
  pull(Bioassay_ID)

tumor_ext_ids <- histology_df %>%
  filter(experimental_strategy == "RNA-Seq",
         sample_type == "Tumor",
         histology == histology_label) %>%
  pull(external_sample_id)

# Subset GSVA and miRNA matrices
pathway_scores <- gsva_scores[, tumor_bioassay_ids, drop=FALSE]
miRNA_expr <- mirna_tpm  [, tumor_ext_ids, drop=FALSE]
```

### Correlate miRNA Expression with Pathway Scores

```{r correlation, message=FALSE, warning=FALSE}
# Load subtype-specific DE miRNA list
sig_DE_miRNA_list <- read.csv(file = de_list_file)

# Optional: exclude specific miRNAs
if (length(exclude_miRNA) > 0) {
  sig_DE_miRNA_list <- sig_DE_miRNA_list %>% filter(!miRNA %in% exclude_miRNA)
}

de_mirnas <- sig_DE_miRNA_list$miRNA
miRNA_expr_sub <- miRNA_expr[rownames(miRNA_expr) %in% de_mirnas, , drop=FALSE]

# Spearman correlation
rho_mat <- cor(t(miRNA_expr_sub), t(pathway_scores), method = "spearman")
rho_mat_filtered <- rho_mat[rowSums(is.na(rho_mat)) != ncol(rho_mat), ]

# p-values (approx)
n <- ncol(miRNA_expr_sub)
t_val <- rho_mat_filtered * sqrt((n-2)/(1 - rho_mat_filtered^2))
pval_mat <- 2 * pt(-abs(t_val), df = n-2)

# Long format with BH adjustment per miRNA
corr_df <- as.data.frame(as.table(rho_mat_filtered)) %>%
  rename(miRNA = Var1, Pathway = Var2, rho = Freq) %>%
  mutate(pval = as.vector(pval_mat)) %>%
  group_by(miRNA) %>%
  mutate(padj = p.adjust(pval, method = "BH")) %>%
  ungroup() %>%
  arrange(miRNA, desc(abs(rho)))
```

### Heatmap: all pairs

```{r draw-heatmap}
rho_mat_all <- corr_df %>%
  select(miRNA, Pathway, rho) %>%
  pivot_wider(names_from = miRNA, values_from = rho) %>%
  column_to_rownames("Pathway") %>%
  as.matrix()

# Ensure excluded miRNAs are not present
if (length(exclude_miRNA) > 0) {
  keep_cols <- setdiff(colnames(rho_mat_all), exclude_miRNA)
  rho_mat_all <- rho_mat_all[, keep_cols, drop = FALSE]
}

# Annotation mapping for "all pairs" view 
sample_anno_all <- tibble(miRNA = colnames(rho_mat_all)) %>%
  left_join(sig_DE_miRNA_list, by = "miRNA")

sample_anno_all$Category  <- factor(sample_anno_all$Category,
                                    levels = c("Paired T vs AN", "Shared", "T vs HN"))
sample_anno_all$direction <- factor(sample_anno_all$direction,
                                    levels = c("up", "down"))

comp_cols <- c("Paired T vs AN" = "#1b9e77",
               "Shared"         = "#d95f02",
               "T vs HN"        = "#7570b3")
dir_cols  <- c(up = "#E41A1C", down = "#377EB8")

# Color mapping for correlation
col_fun <- colorRamp2(c(-1, 0, 1), c("blue", "white", "red"))

ha_all <- HeatmapAnnotation(
  Comparison = sample_anno_all$Category,
  Direction  = sample_anno_all$direction,
  col = list(Comparison = comp_cols, Direction = dir_cols),
  annotation_name_gp = gpar(fontsize = 10),
  annotation_legend_param = list(
    Comparison = list(title = "Comparison", at = levels(sample_anno_all$Category)),
    Direction  = list(title = "Direction",  at = levels(sample_anno_all$direction))
  )
)

ht1 <- Heatmap(rho_mat_all,
               cluster_columns = TRUE, 
               clustering_distance_columns = "euclidean",
               clustering_method_columns = "complete",
               cluster_rows = TRUE,    
               clustering_distance_rows = "euclidean",
               clustering_method_rows = "complete",
               col = col_fun, 
               na_col = "gray", 
               name = "cor",
               show_row_names = FALSE, 
               show_column_names = FALSE,
               top_annotation = ha_all,
               use_raster = TRUE,
               heatmap_legend_param = list(legend_gp = gpar(fontsize = 10),
                                           labels_gp = gpar(fontsize = 10)))
print(ht1)


pdf(file.path(plot_dir, paste0(histology_label, "_heatmap_all.pdf")))
print(ht1)
dev.off()
```

### significant miRNA–pathway pairs

```{r get-signif-corr-df}
signif_corr <- corr_df %>%
  filter(abs(rho) > 0.5, padj < 0.1)

sig_pathway <- unique(signif_corr$Pathway)
sig_mirna   <- unique(signif_corr$miRNA)

corr_df_sig <- corr_df %>%
  filter(Pathway %in% sig_pathway, miRNA %in% sig_mirna)

rho_mat_sig <- corr_df_sig %>%
  select(miRNA, Pathway, rho) %>%
  pivot_wider(names_from = miRNA, values_from = rho) %>%
  column_to_rownames("Pathway") %>%
  as.matrix()
```

### Determining Optimal Clusters

1. Elbow method
2. Silhouette method
3. Gap statistic

```{r run-elbow}
set.seed(123)

# rows (pathways)
fviz_nbclust(rho_mat_sig, # data
             hcut, # clustering algorithm
             method = "wss")  # elbow method

# columns (miRNAs)
fviz_nbclust(t(rho_mat_sig), # data
             hcut, # clustering algorithm
             method = "wss")  # elbow method
```

```{r run-silhouette}
# rows (pathways)
silhouette_pathways <- fviz_nbclust(rho_mat_sig, hcut, method = "silhouette")
print(silhouette_pathways)

# columns (miRNAs)
silhouette_mirnas <- fviz_nbclust(t(rho_mat_sig), hcut, method = "silhouette")
print(silhouette_mirnas)
```

```{r run-gap-stat}
# rows (pathways)
fviz_nbclust(rho_mat_sig, hcut, nboot = 10, method = "gap_stat")

# columns (miRNAs)
fviz_nbclust(t(rho_mat_sig), hcut, nboot = 10, method = "gap_stat")
```

```{r find-optimal-k}
# here silhouette score was used to find optimal k 
sil_dat_row <- silhouette_pathways$data
opt_row_k <- as.integer(sil_dat_row$clusters[which.max(sil_dat_row$y)])
opt_row_k

sil_dat_col <- silhouette_mirnas$data
opt_col_k <- as.integer(sil_dat_col$clusters[which.max(sil_dat_col$y)])
opt_col_k
```

### Heatmap: significant pairs + tables

```{r define-sample-anno}
# Annotation for significant set
sample_anno_sig <- tibble(miRNA = colnames(rho_mat_sig)) %>%
  left_join(sig_DE_miRNA_list, by = "miRNA") %>%
  mutate(
    Category = factor(Category, levels = c("Paired T vs AN", "Shared", "T vs HN")),
    direction = factor(direction, levels = c("up", "down"))
  )

ha_sig <- HeatmapAnnotation(
  Comparison = sample_anno_sig$Category,
  Direction  = sample_anno_sig$direction,
  col = list(Comparison = comp_cols, Direction = dir_cols),
  annotation_name_gp = gpar(fontsize = 10),
  annotation_legend_param = list(
    Comparison = list(title = "Comparison", at = levels(sample_anno_sig$Category)),
    Direction  = list(title = "Direction",  at = levels(sample_anno_sig$direction))
  )
)
```

```{r draw-sig-heatmap, message=FALSE, warning=FALSE}
ht2 <- Heatmap(
  rho_mat_sig,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  row_split = opt_row_k,
  column_split = opt_col_k,
  col = col_fun, na_col = "gray", name = "cor",
  show_row_names = FALSE, show_column_names = FALSE,
  top_annotation = ha_sig,
  use_raster = TRUE,
  heatmap_legend_param = list(legend_gp = gpar(fontsize = 10),
                              labels_gp = gpar(fontsize = 10))
) %>% draw()

row_idx_list <- row_order(ht2)
col_idx_list <- column_order(ht2)
row_idx <- unlist(row_idx_list)
col_idx <- unlist(col_idx_list)

row_cluster_ids <- rep(seq_along(row_idx_list), lengths(row_idx_list))
col_cluster_ids <- rep(seq_along(col_idx_list), lengths(col_idx_list))

write_tsv(
  tibble(Pathway = rownames(rho_mat_sig)[row_idx], row_cluster = row_cluster_ids),
  file.path(results_dir, paste0(histology_label, "_cor_heatmap_pathway_clusters.tsv"))
)

write_tsv(
  tibble(miRNA = colnames(rho_mat_sig)[col_idx], col_cluster = col_cluster_ids),
  file.path(results_dir, paste0(histology_label, "_cor_heatmap_miRNA_clusters.tsv"))
)

# Save figure
pdf(file.path(plot_dir, paste0(histology_label, "_heatmap_sig_cluster.pdf")))
ht2
dev.off()
```

### Export heatmap-ordered matrix with targe genes + annotations

```{r save-heatmap-table, message=FALSE, warning=FALSE}
# Order rows/cols to match the heatmap
rho_mat_ordered <- rho_mat_sig[row_idx, col_idx, drop = FALSE]
miRNA_cols <- colnames(rho_mat_ordered)

mir_lookup  <- setNames(col_cluster_ids, colnames(rho_mat_sig)[col_idx])
path_lookup <- setNames(row_cluster_ids, rownames(rho_mat_sig)[row_idx])

# Target genes mapping
tg_map <- read_tsv(file.path(results_dir, "gobp-gsva-scores-with-target-genes.tsv"), show_col_types = FALSE) %>%
  select(Pathway, TargetGenes = Genes) %>%
  distinct(Pathway, .keep_all = TRUE)

rho_df_chr <- as.data.frame(rho_mat_ordered, stringsAsFactors = FALSE) %>%
  rownames_to_column("Pathway") %>%
  left_join(tg_map, by = "Pathway") %>%
  relocate(TargetGenes, .after = Pathway) %>%
  mutate(
    PathwayCluster = as.character(path_lookup[Pathway])
  ) %>%
  relocate(PathwayCluster, .after = TargetGenes) %>%
  mutate(across(all_of(miRNA_cols), ~ formatC(., digits = 3, format = "f")))

anno_lkp <- sample_anno_sig %>%
  filter(miRNA %in% miRNA_cols) %>%
  arrange(factor(miRNA, levels = miRNA_cols))

comp_row <- tibble(
  Pathway = "Comparison", TargetGenes = "", PathwayCluster = "",
  !!! rlang::set_names(as.list(as.character(anno_lkp$Category)), miRNA_cols)
)
dir_row  <- tibble(
  Pathway = "Direction",  TargetGenes = "", PathwayCluster = "",
  !!! rlang::set_names(as.list(as.character(anno_lkp$direction)), miRNA_cols)
)
mir_row  <- tibble(
  Pathway = "miRNA_cluster", TargetGenes = "", PathwayCluster = "",
  !!! rlang::set_names(as.list(as.character(mir_lookup[miRNA_cols])), miRNA_cols)
)

out_tbl <- bind_rows(comp_row, dir_row, mir_row, rho_df_chr)

write.table(out_tbl,
            file = file.path(results_dir, paste0(histology_label, "_correlation_matrix_heatmap_order.tsv")),
            sep = "\t", quote = FALSE, row.names = FALSE)
```

### Tidy long table

```{r get-tidy-long-table}
miRNA_cluster_df <- tibble(
  miRNA   = colnames(rho_mat_sig)[col_idx],
  Cluster = paste0("C", col_cluster_ids)
)

df_long <- corr_df %>%
  semi_join(
    expand.grid(miRNA = miRNA_cols, Pathway = rownames(rho_mat_sig)[row_idx],
                stringsAsFactors = FALSE),
    by = c("miRNA", "Pathway")
  ) %>%
  left_join(sample_anno_sig %>% select(miRNA, Category, direction), by = "miRNA") %>%
  left_join(miRNA_cluster_df, by = "miRNA") %>%
  left_join(tg_map, by = "Pathway") %>%
  mutate(
    row_rank = match(Pathway, rownames(rho_mat_sig)[row_idx]),
    col_rank = match(miRNA,  colnames(rho_mat_sig)[col_idx])
  ) %>%
  arrange(row_rank, col_rank) %>%
  select(miRNA, Category, direction, Cluster, Pathway, TargetGenes, rho, pval, padj)

write_tsv(df_long, file.path(results_dir, paste0(histology_label, "_miRNA_pathway_correlations_tidy.tsv")))
```

### Session Info

```{r session-info}
sessionInfo()
```


