---
title: 'Assessment of human DMG DE miRNA ortholog expression in B7H3 treated mice'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2026"
---

This script identifies mouse orthologs of human DE miRNAs in DMG versus normal brain by cluster, and assesses their expression patterns in B7H3 treated mice

Load packages 
```{r}
suppressPackageStartupMessages({
  library(rprojroot)
  library(tidyverse)
  library(ggplot2)
  library(gtools)
  library(ggsci)
  library(ggpubr)
})
```

Set dir paths
```{r}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-mirna-expr-in-mouse")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# source plotting theme
source(file.path(root_dir, "figures", "theme.R"))
```

Set file paths
```{r}
# human DMG clustering file
dmg_cluster_file <- file.path(root_dir, "analyses",
                              "mirna-clustering",
                              "results",
                              "DIPG or DMG-de-mirna-cluster-membership-immune-scores.tsv")

# mouse miRNA counts file
mmu_mirna_ct_file <- file.path(data_dir,
                               "30-963755216-miRNA_expression.csv")

# mouse differential expression results
mmu_de_file <- file.path(root_dir, "analyses",
                         "murine-mirna-differential-expression",
                         "results",
                         "mirna-differential-expression-deseq2-b7h3-stop-vs-untreated-by-timepoint.tsv")
```


Wrangle data, and remove species prefix from miRNA IDs to match orthologs
```{r}
# Load hsa DMG miRNA cluster df and filter for cluster 6 miRNAs
dmg_hsa_mir_df <- read_tsv(dmg_cluster_file) %>%
  dplyr::filter(grepl("hsa", miRNA)) %>%
  dplyr::mutate(miRNA = str_remove(miRNA, "hsa-"))

# Load mouse miRNA expression file and determine number of cluster 6 miRNAs with ortholog 
mmu_mirna_ct_df <- read_csv(mmu_mirna_ct_file) %>%
  dplyr::rename("mirna_id" = "...1") %>%
  dplyr::mutate(mirna_id = str_remove(mirna_id, "mmu-"))

# Load merged mouse differential expression results
mmu_de_df <- read_tsv(mmu_de_file) %>%
  dplyr::mutate(mirna_id = str_remove(mirna_id, "mmu-"))
```

Extract DE miRNA groups in mouse data:
1) All DE miRs
2) upregulated D14
3) downregulated D14
4) upregulated D21
5) downregulated D21
```{r}
all_de_mirs <- mmu_de_df %>%
  dplyr::filter(status_Day14 != "Not DE" | status_Day21 != "Not DE") %>%
  pull(mirna_id)

d14_up_mirs <- mmu_de_df %>%
  dplyr::filter(status_Day14 == "B7H3 upregulated") %>%
  pull(mirna_id)

d14_down_mirs <- mmu_de_df %>%
  dplyr::filter(status_Day14 == "B7H3 downregulated") %>%
  pull(mirna_id)

d21_up_mirs <- mmu_de_df %>%
  dplyr::filter(status_Day21 == "B7H3 upregulated") %>%
  pull(mirna_id)

d21_down_mirs <- mmu_de_df %>%
  dplyr::filter(status_Day21 == "B7H3 downregulated") %>%
  pull(mirna_id)
```

Create list of DE miRNA groups
```{r}
de_mir_list <- list("All DE" = all_de_mirs,
                    "D14 Up" = d14_up_mirs,
                    "D14 Down" = d14_down_mirs,
                    "D21 Up" = d21_up_mirs,
                    "D21 Down" = d21_down_mirs)
```

Create data frame to store orthology and expression results by miRNA cluster
```{r}
cluster_summary_df <- tibble(
                             # repeat miR clusters to test for enrichment within each mouse DE group
                             hsa_dmg_cluster = rep(unique(dmg_hsa_mir_df$row_cluster), each = 5),
                             de_mir_group = rep(names(de_mir_list), 8),
                             total_n = NA_integer_,
                             n_mmu_orthologs = NA_integer_,
                             n_expr_orthologs = NA_integer_,
                             n_expr_group = NA_integer_,
                             n_mmu_mir = sum(!grepl("Novel", mmu_de_df$mirna_id)),
                             OR = NA_integer_,
                             OR_p = NA_integer_,
                             OR_CI_lower = NA_integer_,
                             OR_CI_upper = NA_integer_)
```

Loop through rows of `cluster_summary_df` to 
1) identify number of total and expressed mouse orthologs per human DMG miR cluster
2) Assess enrichment of orthologs for miRNAs of different B7H3-associated expression patterns 

```{r}
for (i in 1:nrow(cluster_summary_df)){
  
  # get cluster
  clust <- cluster_summary_df$hsa_dmg_cluster[i]
  
  # get miRNAs in cluster
  cluster_mirs <- dmg_hsa_mir_df %>%
    dplyr::filter(row_cluster == clust) %>%
    # remove hsa prefix from miRNA ID
    dplyr::mutate(miRNA = str_remove(miRNA, "hsa-")) %>%
    pull(miRNA)

  # after removing species prefix, orthologous miRNAs will share the same name between species
  # determine how many hsa DMG cluster miRNAs have orthologs in mice:
  cluster_mirs_w_ortholog <- mmu_mirna_ct_df %>%
    dplyr::filter(mirna_id %in% cluster_mirs) %>%
    pull(mirna_id)

  # get N expressed orthologs from DE df 
  expr_orthologs <- mmu_de_df %>% 
    dplyr::filter(mirna_id %in% cluster_mirs) %>%
    pull(mirna_id)
  
  # append stats to summary df
  cluster_summary_df$total_n[i] <- length(cluster_mirs)
  cluster_summary_df$n_mmu_orthologs[i] <- length(cluster_mirs_w_ortholog)
  cluster_summary_df$n_expr_orthologs[i] <- length(expr_orthologs)
  
  # define current DE group and miRs in group
  de_group <- cluster_summary_df$de_mir_group[i]
  
  de_mirs <- de_mir_list[[de_group]]
  
  # determine number of orthologs in cluster and expression group
  n_ortho_de <- length(intersect(expr_orthologs,
                                 de_mirs))
  
  cluster_summary_df$n_expr_group[i] <- n_ortho_de
  
  # run fisher's exact test to assess for enrichment of orthologs in DE group
  ftest <- fisher.test(matrix(c(n_ortho_de,
                                length(expr_orthologs) - n_ortho_de,
                                length(de_mirs) - n_ortho_de,
                                cluster_summary_df$n_mmu_mir[i] - n_ortho_de - (length(expr_orthologs) - n_ortho_de) - (length(de_mirs) - n_ortho_de)),
                              nrow = 2))
  
  # append FET stats to summary df 
  cluster_summary_df$OR[i] <- ftest$estimate
  cluster_summary_df$OR_p[i] <- ftest$p.value
  cluster_summary_df$OR_CI_lower[i] <- ftest$conf.int[1]
  cluster_summary_df$OR_CI_upper[i] <- ftest$conf.int[2]
 
}
```

Generate df to plot % orthologs identified by cluster

```{r}
ortho_plot_df <- cluster_summary_df %>%
  # retain one row per cluster
  distinct(hsa_dmg_cluster, .keep_all = TRUE) %>%
  dplyr::select(hsa_dmg_cluster,
                total_n,
                n_mmu_orthologs,
                n_expr_orthologs) %>%
  dplyr::rename(`Expressed ortholog` = n_expr_orthologs) %>%
  dplyr::mutate(`Non-expressed ortholog` = n_mmu_orthologs - `Expressed ortholog`,
                `No ortholog` = total_n - n_mmu_orthologs) %>%
  dplyr::select(-n_mmu_orthologs, -total_n) %>%
  pivot_longer(-hsa_dmg_cluster,
               names_to = "category",
               values_to = "n") %>%
  dplyr::mutate(category = fct_relevel(category,
                                       c("Expressed ortholog",
                                         "Non-expressed ortholog",
                                         "No ortholog"))) %>%
  dplyr::mutate(hsa_dmg_cluster = as.factor(hsa_dmg_cluster)) %>%
  dplyr::mutate(hsa_dmg_cluster = fct_relevel(hsa_dmg_cluster,
                                              rev(unique(as.character(hsa_dmg_cluster)))))
```

Plot percent orthologs identified per cluster

```{r}
ortho_plot <- ggplot(ortho_plot_df,
                      aes(x = n,
                          y = hsa_dmg_cluster,
                          fill = category)) +
  geom_bar(stat = "identity",
           color = "black") +
  labs(x = "n miRNAs",
       y = "human DMG DE\nmiRNA cluster") +
  scale_fill_npg() +
  theme_Publication() +
  theme(
    legend.position = "top"
  )

ggsave(file.path(plot_dir, "hsa-dmg-de-mirna-mmu-ortho-barplot.pdf"),
       ortho_plot,
       width = 3.5, height = 4.5)
```

Reorder expr group levels and create columns `perc_text`, `p_text` to show fractions in plots

```{r}
cluster_summary_df <- cluster_summary_df %>% 
  dplyr::mutate(de_mir_group = fct_relevel(de_mir_group,
                                           c("D21 Down",
                                             "D21 Up",
                                             "D14 Down",
                                             "D14 Up",
                                             "All DE"))) %>%
  dplyr::mutate(perc_text = glue::glue("{n_expr_group}/{n_expr_orthologs}")) %>%
  dplyr::mutate(p_text = case_when(
    OR_p < 0.01 ~ "**",
    OR_p < 0.05 ~ "*",
    TRUE ~ ""
  ))
```

Plot enrichment of miRNAs in B7H3-associated expression groups in mouse by cluster

```{r}
enr_plot <- cluster_summary_df %>% 
  # remove clusters with <2 orthologous miRNAs
  dplyr::filter(!hsa_dmg_cluster %in% c(2,7)) %>%
  dplyr::mutate(OR = case_when(
    is.infinite(OR) ~ 1000,
    TRUE ~ OR
  )) %>%
  dplyr::mutate(OR_CI_upper = case_when(
    is.infinite(OR_CI_upper) ~ 100,
    TRUE ~ OR_CI_upper
  )) %>%
  ggplot(aes(x = log10(OR+0.01), y = factor(de_mir_group),
             color = factor(de_mir_group),
             label = p_text)) +
  geom_point(size = 3, show.legend = FALSE) + 
  # add log10-transformed confidence intervals
  geom_errorbar(aes(xmin = log10(OR_CI_lower+0.01), xmax = log10(OR_CI_upper+0.01),
                    color = factor(de_mir_group)), width = 0.2, 
                show.legend = FALSE) +
  geom_text(x = 1.5, hjust = 0, size = 6, fontface = 2,
            show.legend = FALSE) +
  labs(x = "log10-Odds Ratio\n(95% CI)", y = "B7H3 DE miRNA class") + 
  # Add OR threshold line
  geom_vline(xintercept = 0, linetype = "dashed") +
  # facet wrap by variable, place strips to left
  facet_wrap(~hsa_dmg_cluster, scale = "fixed",
             nrow = 8,
             strip.position = "left") +
  expand_limits(y=0) +
  scale_color_manual(values = c("#3C5488FF", "#E64B35FF",
                                "#4DBBD5FF", "#F39B7FFF",
                                "#00A087FF")) +
  theme_Publication() 
```

Plot percent of expressed orthologs in B7H3-associated expression groups by cluster

```{r}
perc_plot <- cluster_summary_df %>%
  dplyr::filter(!hsa_dmg_cluster %in% c(2,7)) %>%
  ggplot(aes(x = n_expr_group/n_expr_orthologs * 100, 
             y = factor(de_mir_group),
             fill = factor(de_mir_group),
             color = factor(de_mir_group),
             label = perc_text)) +
  geom_bar(stat = "identity", color = "black",
           show.legend = FALSE) + 
  # add fractions to the right of plots
  geom_text(x = 60, hjust = 0, size = 4, fontface = 2,
            show.legend = FALSE) +
  labs(x = "% human orthologs\n", y = NULL, fill = NULL) + 
  scale_y_discrete(labels=c("PBTA" = NULL, "gnomAD" = NULL,
                            "PMBB" = NULL)) +
  guides(fill = guide_legend(nrow = 1)) +
  # facet wrap by variable, place strip to left of plot
  facet_wrap(~hsa_dmg_cluster, scale = "fixed",
             nrow = 6,
             strip.position = "left") +
  coord_cartesian(clip = 'off') +
  xlim(c(0, 55)) +
  scale_fill_manual(values = c("#3C5488FF", "#E64B35FF",
                                "#4DBBD5FF", "#F39B7FFF",
                                "#00A087FF")) +
  scale_color_manual(values = c("#3C5488FF", "#E64B35FF",
                                "#4DBBD5FF", "#F39B7FFF",
                                "#00A087FF")) +
  theme_Publication() +
  # remove facet strips
  theme(plot.margin = unit(c(2,5,1,1), "lines"),
        legend.position.inside = c(0.5, 1.07),
        strip.background = element_blank(),
        strip.text.y = element_blank()) +
  theme(strip.placement = "outside")
```

Merge enrichment and percent plots and save

```{r}
merged_plot <- ggarrange(enr_plot, perc_plot,
                         nrow = 1, widths = c(1,0.85),
                         align = "h")

merged_plot

pdf(file.path(plot_dir, 
                 "human-dmg-de-mirna-orthology-mouse-expr-enrichment-plot.pdf"),
       width = 6, height = 8)

merged_plot

dev.off()
```
print session info

```{r}
sessionInfo()
```
