---
title: "Immune Deconvolution Summary"
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2025-07-09"
---

This script summarizes immune deconvolution results as follows:

- Compares immune cell fractions and scores across tumor histologies
- Compares immune cell fractions between tumors and healthy and adjacent normals

### Setup
```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(tidyverse)
library(ggpubr)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "immune-deconvolution")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# File path for immune cell proportions
quantiseq_file <- file.path(results_dir, "quantiseq_output.rds")

xcell_file <- file.path(results_dir, "xcell_output.rds")

# File path for histology
histology_file <- file.path(root_dir, "analyses", "histology-preprocessing", "results", "histologies.tsv")
```

Wrangle data

```{r}
histologies <- read_tsv(histology_file)
```

Load quantiseq deconvolution output

```{r}
quantiseq_res <- readRDS(quantiseq_file) %>%
  # define groups for plotting
  dplyr::mutate(group = case_when(
    grepl("Adjacent", sample_type) ~ "Adjacent N",
    sample_type == "Normal" ~ "Healthy N",
    TRUE ~ "Tumor"
  ))
```

Identify paired samples 

```{r}
paired_samples <- histologies %>%
  dplyr::filter(experimental_strategy == "RNA-Seq") %>%
  dplyr::count(sample_id) %>% 
  dplyr::filter(n == 2) %>%
  pull(sample_id)
```

Plot quantiseq-derived cell proportions in tumors by histology 

```{r}
hist_contrasts <- list(c("DIPG or DMG", "EPN"),
                                        c("DIPG or DMG", "LGG"),
                                        c("DIPG or DMG", "MB"))

quantiseq_res %>%
  dplyr::filter(sample_type == "Tumor",
                histology %in% c("DIPG or DMG", "EPN", "MB", "LGG"),
                cell_type != "uncharacterized cell") %>%
  
  ggplot(aes(x = histology, y = fraction)) + 
  geom_jitter(aes(fill = histology), shape = 21, 
              width = 0.2, size = 2.5, alpha = 0.85,
              show.legend = FALSE) +
  geom_boxplot(alpha = 0.05, outlier.shape = NA) +
  stat_compare_means(method = "wilcox",
                     comparisons = hist_contrasts,
                     method.args = list(alternative = "two.sided"),
                       label = "p.signif") +
  labs(x = NULL, y = "Cell fraction") +
  facet_wrap(~cell_type, nrow = 2, scales = "free_y",
             labeller = labeller(cell_type = label_wrap_gen(18))) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.25))) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(plot_dir,
                   glue::glue("quantiseq-cell-proportions-byHist.pdf")),
         width = 13, height = 6)
```
compute and plot M2:M1 macrophage ratio across tumor histologies

```{r}
macrophage_quantiseq_df <- quantiseq_res %>%
  dplyr::filter(sample_type == "Tumor",
                histology %in% c("DIPG or DMG", "EPN", "MB", "LGG"),
                grepl("Macrophage", cell_type)) %>%
  pivot_wider(names_from = cell_type,
              values_from = fraction,
              id_cols = c(Bioassay_ID, histology)) %>%
  dplyr::mutate(m2_m1_ratio = `Macrophage M2`/`Macrophage M1`) %>%
  dplyr::mutate(m2_m1_ratio = case_when(
    is.infinite(m2_m1_ratio) ~ 5,
    is.nan(m2_m1_ratio) ~ 0,
    TRUE ~ m2_m1_ratio
  ))

macrophage_quantiseq_df %>%
  
  ggplot(aes(x = histology, y = m2_m1_ratio)) + 
  geom_jitter(aes(fill = histology), shape = 21, 
              width = 0.2, size = 2.5, alpha = 0.85,
              show.legend = FALSE) +
  geom_boxplot(alpha = 0.05, outlier.shape = NA) +
  stat_compare_means(method = "wilcox",
                     comparisons = hist_contrasts,
                     method.args = list(alternative = "two.sided"),
                       label = "p.signif") +
  labs(x = NULL, y = "M2:M1 ratio") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.25))) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(plot_dir,
                   glue::glue("quantiseq-m2-m1-macrophage-ratio-byHist.pdf")),
         width = 4, height = 3)
```

Loop through histologies and plot quantiseq cell fractions in tumor vs. normal

```{r}
groups <- c("DIPG or DMG", "EPN", "LGG", "MB")

for (hist in groups){
  
  hist_plot_dir <- file.path(plot_dir, hist)
  
  if (!dir.exists(hist_plot_dir)){
    dir.create(hist_plot_dir)
  }
  
  # subset for histology
  quantiseq_group <- quantiseq_res %>%
    dplyr::filter(histology == hist | sample_type == "Normal",
                cell_type != "uncharacterized cell")

# plot tumor vs adj and healthy normal
  quantiseq_group %>%
    dplyr::mutate(group = fct_relevel(group,
                                      c("Healthy N",
                                        "Adjacent N",
                                        "Tumor"))) %>%
    dplyr::arrange(group) %>%
    
    ggplot(aes(x = factor(group), y = fraction)) + 
    geom_jitter(aes(fill = group), shape = 21, 
                width = 0.2, size = 2.5, alpha = 0.85,
                show.legend = FALSE) +
    geom_boxplot(alpha = 0.05, outlier.shape = NA) +
    stat_compare_means(method = "wilcox",
                       comparisons = list(c("Tumor", "Healthy N"),
                                          c("Tumor", "Adjacent N")),
                       method.args = list(alternative = "two.sided"),
                       label = "p.signif") +
    labs(x = NULL, y = "Cell fraction") +
    facet_wrap(~cell_type, nrow = 2, scales = "free_y",
               labeller = labeller(cell_type = label_wrap_gen(18))) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) +
    theme_Publication() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
    
  ggsave(file.path(hist_plot_dir,
                   glue::glue("{hist}-cell-proportions-tumor-v-adj-healthyN-quantiseq.pdf")),
         width = 15, height = 6)
    
  # plot paired tumor-adj samples
  if (hist != "LGG"){
  
    quantiseq_group %>%
      dplyr::filter(!is.na(histology),
                    sample_id %in% paired_samples) %>%
      dplyr::arrange(cell_type, sample_id, group) %>%
      
      ggplot(aes(x = group, y = fraction)) + 
      geom_jitter(aes(fill = group), shape = 21, 
                  width = 0.2, size = 2.5, alpha = 0.85,
                  show.legend = FALSE) +
      geom_boxplot(alpha = 0.05, outlier.shape = NA) +
      stat_compare_means(method = "wilcox",
                         comparisons = list(c("Tumor", "Adjacent N")),
                         method.args = list(alternative = "two.sided"),
                       label = "p.signif") +
      labs(x = NULL, y = "Cell fraction") +
      facet_wrap(~cell_type, nrow = 2, scales = "free_y",
                 labeller = labeller(cell_type = label_wrap_gen(18))) +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) +
      theme_Publication() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    
    ggsave(file.path(hist_plot_dir,
                 glue::glue("{hist}-cell-proportions-tumor-v-adjN-quantiseq.pdf")),
         width = 12, height = 5)
      
  }
  
}
```

Load Xcell immune deconvolution output

```{r}
xcell_res <- readRDS(xcell_file) %>%
  dplyr::mutate(group = case_when(
    grepl("Adjacent", sample_type) ~ "Adjacent N",
    sample_type == "Normal" ~ "Healthy N",
    TRUE ~ "Tumor"
  )) %>%
  # define groups of cell types for plotting
  dplyr::mutate(cell_group = case_when(
      grepl("B cell", cell_type) ~ "B cell",
      grepl("T cell", cell_type) ~ "T cell",
      grepl("Macrophage|Monocyte|Myeloid|Neutrophil|Mast|dendritic", cell_type) ~ "Innate immune cell",
      grepl("score", cell_type) ~ "Score",
      TRUE ~ "Other"
    ))
```

Plot tumor cell enrichment scores by histology and cell group

```{r}
for (cell in unique(xcell_res$cell_group)){
  
  ncells <- length(unique(xcell_res$cell_type[xcell_res$cell_group == cell]))
  
  plot_width <- ifelse(cell == "B cell", 6,
                     ifelse(cell == "Innate immune cell", 9,
                            ifelse(cell == "T cell", 12,
                                   ifelse(cell == "Other", 8, 3))))
  
  xcell_res %>%
    dplyr::filter(sample_type == "Tumor",
                  histology %in% c("DIPG or DMG", "EPN", "LGG", "MB"),
                  cell_group == cell) %>%
    
    ggplot(aes(x = histology, y = fraction)) + 
    geom_jitter(aes(fill = histology), shape = 21, 
                width = 0.2, size = 2.5, alpha = 0.85,
                show.legend = FALSE) +
    geom_boxplot(alpha = 0.05, outlier.shape = NA) +
    stat_compare_means(method = "wilcox",
                       comparisons = hist_contrasts,
                       method.args = list(alternative = "two.sided"),
                       label = "p.signif") +
    labs(x = NULL, y = "Cell enrichment score") +
    facet_wrap(~cell_type, nrow = 3, scales = "free_y",
               labeller = labeller(cell_type = label_wrap_gen(18))) +
    theme_Publication() +
        scale_y_continuous(expand = expansion(mult = c(0.05, 0.25))) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  ggsave(file.path(plot_dir,
                   glue::glue("{cell}-proportions-byHist-xcell.pdf")),
         width = plot_width, height = 7)
    
}
```

Compute and plot M2:M1 macrophage ratio by histology

```{r}
macrophage_xcell_df <- xcell_res %>%
  dplyr::filter(sample_type == "Tumor",
                histology %in% c("DIPG or DMG", "EPN", "MB", "LGG"),
                grepl("Macrophage M", cell_type)) %>%
  dplyr::mutate(fraction = case_when(
    fraction < 1e-5 ~ 0,
    TRUE ~ fraction
  )) %>%
  pivot_wider(names_from = cell_type,
              values_from = fraction,
              id_cols = c(Bioassay_ID, histology)) %>%
  dplyr::mutate(m2_m1_ratio = `Macrophage M2`/`Macrophage M1`) %>%
  dplyr::mutate(m2_m1_ratio = case_when(
    is.infinite(m2_m1_ratio) ~ 20,
    is.nan(m2_m1_ratio) ~ 0,
    TRUE ~ m2_m1_ratio
  ))

macrophage_xcell_df %>%
  
  ggplot(aes(x = histology, y = m2_m1_ratio)) + 
  geom_jitter(aes(fill = histology), shape = 21, 
              width = 0.2, size = 2.5, alpha = 0.85,
              show.legend = FALSE) +
  geom_boxplot(alpha = 0.05, outlier.shape = NA) +
  stat_compare_means(method = "wilcox",
                     comparisons = hist_contrasts,
                     method.args = list(alternative = "two.sided"),
                       label = "p.signif") +
  labs(x = NULL, y = "M2:M1 ratio") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.25))) +
  theme_Publication() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(plot_dir,
                   glue::glue("xcell-m2-m1-macrophage-ratio-byHist.pdf")),
         width = 4, height = 3)
```

Loop through histologies and plot tumor vs. normal cell enrichment scores 

```{r}
for (hist in groups){
  
  hist_plot_dir <- file.path(plot_dir, hist)
  
  if (!dir.exists(hist_plot_dir)){
    dir.create(hist_plot_dir)
  }
  
  # subset for histology
  xcell_group <- xcell_res %>%
    dplyr::filter(histology == hist | sample_type == "Normal")
  
  # remove cell types with low proportions
  mean_cell_proportions <- xcell_group %>%
    group_by(cell_type) %>%
    summarise(mean_proportion = mean(fraction, na.rm = TRUE)) %>%
    arrange(mean_proportion)
  
  cell_types_retain <- mean_cell_proportions %>%
    dplyr::filter(mean_proportion >= 1e-5) %>%
    pull(cell_type)
  
  xcell_group <- xcell_group %>% 
    dplyr::filter(cell_type %in% cell_types_retain)
  
  # loop through cell groups for plotting
  for (cell in unique(xcell_group$cell_group)){
    
    # determine plot widths based on n cell types
    ncells <- length(unique(xcell_group$cell_type[xcell_group$cell_group == cell]))
  
    plot_width <- ifelse(cell == "B cell", 6,
                       ifelse(cell == "Innate immune cell", 9,
                              ifelse(cell == "T cell", 12,
                                     ifelse(cell == "Other", 8, 3))))
  
    # plot
    xcell_group %>%
      dplyr::filter(cell_group == cell) %>%
      dplyr::mutate(group = fct_relevel(group,
                                      c("Healthy N",
                                        "Adjacent N",
                                        "Tumor"))) %>%
    dplyr::arrange(group) %>%
      
      ggplot(aes(x = factor(group), y = fraction)) + 
      geom_jitter(aes(fill = group), shape = 21, 
                  width = 0.2, size = 2.5, alpha = 0.85,
                  show.legend = FALSE) +
      geom_boxplot(alpha = 0.05, outlier.shape = NA) +
      stat_compare_means(method = "wilcox",
                         comparisons = list(c("Tumor", "Healthy N"),
                                            c("Tumor", "Adjacent N")),
                         method.args = list(alternative = "two.sided"),
                       label = "p.signif") +
      labs(x = NULL, y = "Cell enrichment score") +
      facet_wrap(~cell_type, nrow = 3, scales = "free_y",
                 labeller = labeller(cell_type = label_wrap_gen(18))) +
      theme_Publication() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      scale_y_continuous(expand = expansion(mult = c(0.05, 0.2)))
    
    # save to output
    ggsave(file.path(hist_plot_dir,
                     glue::glue("{hist}-{cell}-proportions-tumor-v-adj-healthyN-xcell.pdf")),
           width = plot_width, height = 7)
    
    # plot matched T/N samples only
    if (hist != "LGG"){
    
      xcell_group %>%
        dplyr::filter(!is.na(histology),
                      sample_id %in% paired_samples,
                      cell_group == cell) %>%
        dplyr::arrange(cell_type, sample_id, group) %>%
        
        ggplot(aes(x = group, y = fraction)) + 
        geom_jitter(aes(fill = group), shape = 21, 
                    width = 0.2, size = 2.5, alpha = 0.85,
                    show.legend = FALSE) +
        geom_boxplot(alpha = 0.05, outlier.shape = NA) +
        stat_compare_means(method = "wilcox",
                           comparisons = list(c("Tumor", "Adjacent N")),
                           method.args = list(alternative = "two.sided"),
                       label = "p.signif") +
        labs(x = NULL, y = "Cell enrichment score") +
        facet_wrap(~cell_type, nrow = 3, scales = "free_y",
                   labeller = labeller(cell_type = label_wrap_gen(18))) +
        theme_Publication() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        scale_y_continuous(expand = expansion(mult = c(0.05, 0.2)))
      
      ggsave(file.path(hist_plot_dir,
                   glue::glue("{hist}-{cell}-proportions-tumor-v-adjN-xcell.pdf")),
           width = plot_width, height = 7)
      
    }
    
  }
  
}
```

Session info
```{r}
sessionInfo()
```
