---
title: "Calculate miRNA expression-quantiseq immune cell fraction correlations"
output:
  html_document:
    toc: true
    toc_float: true
author: Ryan Corbett
date: "2025-07-09"
---

### Setup
```{r setup, message=FALSE, warning=FALSE}
# Load libraries
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(ggpubr)
#  library(ggrepel)
 # library(patchwork)
 # library(matrixStats)
  library(ggsci)
})

# Define directories
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "immune-deconvolution")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# File path for counts
mirna_cts_file <- file.path(root_dir, "analyses", 
                            "human-mirna-expression", "results", 
                            "merged-miRNA-expression.tsv")

# File path for immune cell proportions
quantiseq_file <- file.path(results_dir, "quantiseq_output.rds")

# File path for histology
histology_file <- file.path(root_dir, "analyses", 
                            "histology-preprocessing", 
                            "results", "histologies.tsv")
```

### Data Wrangling
```{r data-wrangle}
# Load and filter miRNA counts
mirna_cts <- read_tsv(mirna_cts_file) %>%
  column_to_rownames("miRNA")
```

Rename healthy normal samples

```{r rename-sample}
names(mirna_cts)[1] <- "4-1958-CC1"
names(mirna_cts)[2] <- "2-1578-CC1"
names(mirna_cts)[3] <- "5-1234-left-pons"
names(mirna_cts)[4] <- "1-1855-CC1"
names(mirna_cts)[5] <- "3-1890-CC1"
```

Remove infiltrated pons samples

```{r remove-tumor-infiltrated-sample}
# Remove 5-1234-left-pons showing significant tumor infiltration
mirna_cts <- mirna_cts[, !(colnames(mirna_cts) == "5-1234-left-pons")]
```

Fix tumor-normal mismatch

```{r fix-tumor-normal-swap}
# Get current column names
current_colnames <- colnames(mirna_cts)

# Find the indices of the two samples
idx_T <- which(current_colnames == "2058-T")
idx_N <- which(current_colnames == "2058-N")

colnames(mirna_cts)[c(idx_T, idx_N)] <- colnames(mirna_cts)[c(idx_N, idx_T)]
```

Load histológies file

```{r}
hist <- read_tsv(histology_file)
```

Calculate miRNA TPM

```{r}
# Assume fixed miRNA length of 22 bases (0.022 kb)
mirna_length_kb <- 22 / 1000  

# Function to compute TPMs
calculate_tpm <- function(count_matrix, gene_length_kb) {
  rpk <- count_matrix / gene_length_kb
  tpm <- sweep(rpk, 2, colSums(rpk), FUN = "/") * 1e6
  return(tpm)
}

# calculate tpm
mirna_tpm <- calculate_tpm(mirna_cts, gene_length_kb = mirna_length_kb)

# save TPM file to output
saveRDS(mirna_tpm,
          file.path(results_dir,
                    "mirna-tpm.rds"))
```

Define samples of interest (DIPG/DMG and MB samples) and filter TPM matrix

```{r}
samples_of_interest <- hist %>%
  dplyr::filter(experimental_strategy == "miRNA-Seq",
                histology %in% c("DIPG or DMG", "MB"),
                external_sample_id != "5-1234-left-pons",
                !grepl("2237", external_sample_id)) %>%
  pull(external_sample_id)

mirna_tpm <- mirna_tpm[,samples_of_interest]
```

Load quantiseq results and append bioassay IDs

```{r}
quantiseq_res <- readRDS(quantiseq_file) %>%
  left_join(hist %>% dplyr::select(external_sample_id,
                                   Bioassay_ID))
```

Loop through histologies to calculate miRNA expr-cell fraction correlations

```{r}
groups <- c("DIPG or DMG", "MB")

cor_list <- list()

for (group in groups){
  
  plot_dir <- file.path(analysis_dir, "plots",
                           group)
  
  # define samples of interest
  group_samples <- hist %>% 
    dplyr::filter(histology == group,
                  !grepl("Adjacent", sample_type),
                  external_sample_id != "5-1234-left-pons",
                !grepl("2237", external_sample_id)) %>%
    distinct(external_sample_id) %>%
    pull(external_sample_id)
  
  # Load DE miRNA results (healthy N and adjacent N references)
  mirna_de_healthyN <- read_csv(file.path(root_dir, "analyses",
                                          "human-mirna-expression",
                                          "results",
                                          glue::glue("DESeq2_{group}_vs_healthyNormal_sig.csv")))
  
  mirna_de_adjN <- read_csv(file.path(root_dir, "analyses",
                                          "human-mirna-expression",
                                          "results",
                                          glue::glue("DESeq2_{group}_paired_sig.csv")))
  
  # identify common DE mirnas
  common_de_mirnas <- intersect(mirna_de_healthyN$miRNA,
                                mirna_de_adjN$miRNA)
  
  # filter tpm matrix for DE mirnas and samples in group
  de_mirna_tpm <- mirna_tpm[common_de_mirnas, group_samples]
  
  # convert tpm matrix to data frame
  de_mirna_df <- de_mirna_tpm %>%
    t() %>%
    as.data.frame() %>%
    rownames_to_column("external_sample_id") %>%
    pivot_longer(-external_sample_id,
                 names_to = "miRNA",
                 values_to = "tpm") %>%
    # append cell fractions
    full_join(quantiseq_res %>% 
                dplyr::filter(external_sample_id %in% colnames(de_mirna_tpm)) %>%
                dplyr::select(external_sample_id,
                              cell_type,
                              sample_type,
                              fraction)) %>%
    # calculate log2 tpm
    dplyr::mutate(log2_tpm = log2(tpm+0.01)) %>%
    dplyr::filter(!is.na(log2_tpm))
  
  # loop through miRNAs and generate tpm-cell score scatter plots
  for (mirna in unique(de_mirna_df$miRNA)) {
    
  # generate scatter plot
  de_mirna_df %>%
      dplyr::filter(miRNA == mirna,
                    !is.na(cell_type),
                    cell_type != "uncharacterized cell") %>%
      
      ggscatter(x = "log2_tpm", 
              y= "fraction", 
              color = "sample_type",
              add = "reg.line", 
              conf.int = TRUE, 
              cor.coef = TRUE, 
              cor.method = "pearson",
              add.params = list(color = "black",
                                fill = NA),
              ticks = TRUE,
              alpha = 0.6) + 
      facet_wrap("cell_type", labeller = labeller(cell_type = label_wrap_gen(width = 18)), 
                 nrow  = 3,
                 scales = "free_y") +
      theme_Publication() + 
      labs(x = glue::glue("{mirna} log2-TPM"),
           y = "Cell fraction") + 
      theme(strip.text = element_text(size = 13))  # Adjust the size here
    
    # save to output
    ggsave(file.path(plot_dir,
                   glue::glue("{mirna}-tpm-vs-quantiseq-cell-proportions-{group}.pdf")),
         width = 14, height = 8)
    
  }
    
    # generate correlation data frame and append empty pearson correlation and p-value columns
    cor_list[[group]] <- tibble(miRNA = unique(de_mirna_df$miRNA)) %>%
      left_join(de_mirna_df %>% dplyr::select(miRNA,
                                              cell_type)) %>%
      dplyr::filter(!is.na(cell_type)) %>%
      distinct() %>%
      dplyr::mutate(pearson_r = NA_integer_,
                    pearson_pvalue = NA_integer_)
    
    # run correlation analysis on each miRNA-cell type pair
    for (i in 1:nrow(cor_list[[group]])){
      
      cor_test <- cor.test(de_mirna_df$log2_tpm[de_mirna_df$miRNA == cor_list[[group]]$miRNA[i] & de_mirna_df$cell_type == cor_list[[group]]$cell_type[i]],
                           de_mirna_df$fraction[de_mirna_df$miRNA == cor_list[[group]]$miRNA[i] & de_mirna_df$cell_type == cor_list[[group]]$cell_type[i]],
                           method = "pearson")
      
      cor_list[[group]]$pearson_r[i] <- cor_test$estimate
      cor_list[[group]]$pearson_pvalue[i] <- cor_test$p.value
      
    }
    
  # write to output
  write_tsv(cor_list[[group]],
        file.path(results_dir, glue::glue("{group}-de-mirna-quantiseq-fraction-correlations.tsv")))
  
}


```

print session info
```{r}
sessionInfo()
```