---
title: 'Human miRNA differential expression in DIPG/DMG versus normal brain'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2025"
---

This script performs differential expression analysis of human miRNAs in DIPG/DMG samples versus normal pons

Load packages

```{r load packages}
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(EnhancedVolcano)
```

Set directory paths

```{r dir paths}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-mirna-differential-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")
```

Set file paths 

```{r file paths}
mirna_cts_file <- file.path(data_dir,
                          "30-931106737-miRNA_expression.csv")
```

Wrangle data

```{r wrangle data}
mirna_cts <- read_csv(mirna_cts_file) %>%
  column_to_rownames("...1")
```

Create sample metaata df

```{r metadata}
metadata_df <- tibble(id = names(mirna_cts)) %>%
  dplyr::mutate(group = case_when(
    grepl("left-pons", id) ~ "normal pons",
    TRUE ~ "DIPG_DMG"
  )) %>%
  as.data.frame()
```

Construct DESeqDataSet Object

```{r create dds}
dds <- DESeqDataSetFromMatrix(countData = mirna_cts, 
                              colData = metadata_df, 
                              design = ~group)
```

Plot miRNA expression PCA

```{r pca}
# transform raw counts
vsdata <- varianceStabilizingTransformation(dds)

pdf(file.path(plot_dir,
              "mirna-pca-plot.pdf"),
    width = 5, height = 3)

plotPCA(vsdata, intgroup="group")

dev.off()
```

Run DESeq2 function

```{r deseq2}
dds <- DESeq(dds)
```

Write results to variable

```{r results}
# set contrast as dipg/dmg versus normal
res <- results(dds, contrast = c("group", "DIPG_DMG", "normal pons"))

summary(res)
```

Generate differential expression volcano plot

```{r volcanoplot}
volc_plot <- EnhancedVolcano(res,
                             lab = rownames(res), 
                             x = "log2FoldChange",
                             y = "pvalue",
                             title = "DIPG/DMG versus normal pons",
                             subtitle = NULL,
                             caption = NULL,
                             pCutoff = max(res$pvalue[res$padj < 0.05], na.rm = TRUE),
                             FCcutoff = 2,
                             pointSize = 4,
                             labSize = 3,
                             typeConnectors = "closed",
                             drawConnectors = TRUE,
                             widthConnectors = 1,
                             colConnectors = 'black',
                             max.overlaps = 8)

pdf(file.path(plot_dir,
              "de-mirna-dipg-dmg-versus-normal-volcano-plot.pdf"),
    width = 6, height = 6)

volc_plot

dev.off()
```

Print DIPG/DMG upregulated miRNAs

```{r}
res <- data.frame(results(dds))

res %>%
  dplyr::filter(log2FoldChange > 0,
                padj < 0.05)
```

Print DIPG/DMG downregulated miRNAs

```{r}
res %>%
  dplyr::filter(log2FoldChange < 0,
                padj < 0.05)
```

Plot counts of differentially expression miRNAs

```{r plot cts}
de_mirnas <- res %>%
  arrange(pvalue) %>%
  dplyr::filter(padj < 0.05) %>%
  rownames_to_column("mirna_id") %>%
  pull(mirna_id)

for (mirna in de_mirnas){
  
  pdf(file.path(plot_dir,
                glue::glue("{mirna}_cts_dipg_dmg_versus_normal.pdf")),
      width = 5, height = 5)
  
  plotCounts(dds,
             gene = mirna,
             intgroup = "group")
  
  dev.off()
  
}
```

Save differential expression results

```{r write output}
res %>% 
  rownames_to_column("mirna_id") %>%
  write_tsv(file.path(results_dir,
                      "mirna-differential-expr-deseq2-dipg-dmg-versus-normal.tsv"))
```

Print session info

```{r session info}
sessionInfo()
```