---
title: "Neuronal & Synaptic Gene Dysregulation Across Pediatric Brain Tumors"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
author: Bicna Song
date: "`r Sys.Date()`"
---

This analysis evaluates expression changes of neuronal gene programs across pediatric brain tumors by leveraging previously generated differential expression results from DESeq2.

We analyze the following comparison outputs:

- **Tumor vs Tumor-Adjacent Normal**
- **Tumor vs Healthy Normal Brain (B1–B5)**

Tumor groups included:

- DIPG/DMG  
- MB  
- EPN  
- LGG (healthy-normal only)

---

### Setup

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(rprojroot)
library(ggplot2)
library(ggpubr)
library(Biobase)

root_dir <- find_root(has_dir(".git"))
analysis_dir <- file.path(root_dir, "analyses", "human-neuronal-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

source(file.path(root_dir, "figures", "theme.R"))
``` 

### Define Neuronal & Synaptic Gene Sets

```{r define-gene-sets}
gene_sets <- list(
  synaptic_vesicle_trafficking = c(
    "STX1A","STX1B","SNAP25","VAMP2","SYT1","SYT2","SYT7","CPLX1","CPLX2",
    "RAB3A","RAB3B","RAB3C","RAB27A","UNC13A","UNC13B","STXBP1"),
  neurotransmitter_receptors = c(
    paste0("GRIA",1:4),
    "GRIN1", paste0("GRIN2",c("A","B","C","D")), paste0("GRIN3",c("A","B")),
    paste0("GRIK",1:5),
    paste0("GABRA",1:6), paste0("GABRB",1:3), paste0("GABRG",1:3),
    "GABRD","GABBR1","GABBR2"),
  postsynaptic_scaffolding = c(
    "DLG4", paste0("DLG",1:3),
    paste0("SHANK",1:3), paste0("HOMER",1:3),
    paste0("DLGAP",1:4)),
  cell_adhesion = c(
    paste0("NRXN",1:3),
    paste0("NLGN",1:4),
    "CDH2",
    paste0("PCDH", c("A1","A2","A3","B2","GA3","GB4")),
    paste0("LRRTM",1:4)),
  neuromodulatory = c(
    paste0("DRD",1:5),
    paste0("HTR1",c("A","B","D","E","F")), paste0("HTR2",c("A","B","C")),
    paste0("HTR3",c("A","B","C","D","E")),
    paste0("HTR",4:7),
    paste0("CHRNA",1:10), paste0("CHRNB",1:4), paste0("CHRM",1:5),
    "NPY","PENK","TAC1","VIP"),
  glutamatergic = c(
    paste0("GRIA",1:4), "GRIN1", paste0("GRIN",2:3), paste0("GRIK",1:5),
    paste0("SLC1A",1:3), "GLUL","GLS","GLS2"),
  gabaergic = c(
    paste0("GABRA",1:6), paste0("GABRB",1:3), paste0("GABRG",1:3),
    "GABRD","GABBR1","GABBR2","SLC6A1","SLC6A11","GAD1","GAD2","ABAT"),
  cholinergic = c(
    paste0("CHRNA",1:10), paste0("CHRNB",1:4), paste0("CHRM",1:5),
    "SLC5A7","CHAT","ACHE"),
  serotonergic = c(
    paste0("HTR",1:7), "SLC6A4","TPH1","TPH2","MAOA","MAOB"),
  purinergic = c(
    paste0("P2RX",1:7),
    paste0("P2RY",1:14),
    paste0("ENTPD",1:3),
    "NT5E"),
  adrenergic = c("ADRA1A","ADRA1B","ADRA1D","ADRA2A","ADRA2B","ADRA2C",
                 paste0("ADRB",1:3),"SLC6A2","DBH","MAOA","MAOB","COMT")
)

category_df <- enframe(gene_sets, name = "category", value = "gene") %>%
  unnest(cols = "gene") %>%
  mutate(gene = toupper(gene))

category_df$unique <- isUnique(category_df$gene)
```

### Load DESeq2 Results

```{r load-deseq2}
# We will use the significant DE gene output files:
de_files <- c(
  "DESeq2_DIPG or DMG_paired_sig.csv",
  "DESeq2_DIPG or DMG_vs_healthyNormal_sig.csv",
  "DESeq2_EPN_paired_sig.csv",
  "DESeq2_EPN_vs_healthyNormal_sig.csv",
  "DESeq2_LGG_vs_healthyNormal_sig.csv",
  "DESeq2_MB_paired_sig.csv",
  "DESeq2_MB_vs_healthyNormal_sig.csv"
)

de_files <- file.path(root_dir, "analyses", "human-rna-expression", "results", de_files)
```

```{r extract-tumor-group}
# A helper function identifies the tumor group from the filename:
extract_group <- function(filename) {
  if (grepl("DIPG|DMG", filename, ignore.case = TRUE)) return("DIPG_or_DMG")
  if (grepl("MB", filename, ignore.case = TRUE)) return("MB")
  if (grepl("EPN", filename, ignore.case = TRUE)) return("EPN")
  if (grepl("LGG", filename, ignore.case = TRUE)) return("LGG")
  return("Unknown")
}
```

### Generate Neuronal/Synaptic Gene Plots for Each Tumor Type

```{r generate-plots}
for (f in de_files) {
  
  if (!file.exists(f)) next
  
  tumor_group <- extract_group(f)
  df <- read_csv(f) %>%
    mutate(gene = toupper(gene_symbol))
  
  merged <- inner_join(df, category_df, by = "gene")
  
  if (nrow(merged) == 0) {
    message("No neuronal/synaptic genes found in: ", f)
    next
  }
  
  merged <- merged %>%
    mutate(
      fc_direction = if_else(log2FoldChange < 0, "Down", "Up"),
      gene = reorder(gene, log2FoldChange)
    )
  
  plot_title <- paste0("Neuronal & Synaptic Gene Dysregulation in ", tumor_group)
  
  p <- ggplot(
        merged,
        aes(
          x = gene,
          y = log2FoldChange,
          size = -log10(padj),
          color = fc_direction
        )
      ) +
    geom_point(alpha = 0.85) +
    facet_wrap(~ category, scales = "free_x", nrow = 2) +
    theme_Publication() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 6),  # NEW: smaller font
      strip.text = element_text(size = 10, face = "bold")
    ) +
    labs(
      title = plot_title,
      subtitle = basename(f),
      x = "Gene",
      y = "log2FC (Tumor vs Normal)",
      size = "-log10(padj)",
      color = "Direction"
    ) +
    scale_color_manual(values = c("Down" = "red3", "Up" = "green4"))
  
  # --- Clean output filename ---
  base <- basename(f)
  
  # Extract comparison (paired or vs_healthyNormal)
  comparison <- case_when(
    grepl("paired", base, ignore.case = TRUE) ~ "paired",
    grepl("vs_healthyNormal", base, ignore.case = TRUE) ~ "healthyNormal",
    TRUE ~ "comparison"
  )
  
  # Build new clean filename
  clean_name <- paste0("neuronal_", tumor_group, "_", comparison, ".pdf")

  out_file <- file.path(plot_dir, clean_name)
  
  ggsave(out_file, p, width = 20, height = 10)
  message("Saved: ", out_file)
}
```

### Session Info

```{r session-info}
sessionInfo()
```

