---
title: "Neuronal & Synaptic Gene Dysregulation Across Pediatric Brain Tumors"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
author: Bicna Song
date: "`r Sys.Date()`"
---

This analysis evaluates expression changes of neuronal gene programs across pediatric brain tumors by leveraging previously generated differential expression results from DESeq2.

We analyze the following comparison outputs:

- **Tumor vs Tumor-Adjacent Normal**
- **Tumor vs Healthy Normal Brain (B1–B5)**

Tumor groups included:

- DIPG/DMG  
- MB  
- EPN  
- LGG (healthy-normal only)

---

### Setup

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(rprojroot)
library(ggplot2)
library(ggpubr)
library(Biobase)
library(ggforce)
library(cowplot)

root_dir <- find_root(has_dir(".git"))
analysis_dir <- file.path(root_dir, "analyses", "human-neuronal-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

source(file.path(root_dir, "figures", "theme.R"))
``` 

### Define Neuronal & Synaptic Gene Sets

```{r define-gene-sets}
gene_sets <- list(
  synaptic_vesicle_trafficking = c(
    "STX1A","STX1B","SNAP25","VAMP2","SYT1","SYT2","SYT7","CPLX1","CPLX2",
    "RAB3A","RAB3B","RAB3C","RAB27A","UNC13A","UNC13B","STXBP1"),
  neurotransmitter_receptors = c(
    paste0("GRIA",1:4),
    "GRIN1", paste0("GRIN2",c("A","B","C","D")), paste0("GRIN3",c("A","B")),
    paste0("GRIK",1:5),
    paste0("GABRA",1:6), paste0("GABRB",1:3), paste0("GABRG",1:3),
    "GABRD","GABBR1","GABBR2"),
  postsynaptic_scaffolding = c(
    "DLG4", paste0("DLG",1:3),
    paste0("SHANK",1:3), paste0("HOMER",1:3),
    paste0("DLGAP",1:4)),
  cell_adhesion = c(
    paste0("NRXN",1:3),
    paste0("NLGN",1:4),
    "CDH2",
    paste0("PCDH", c("A1","A2","A3","B2","GA3","GB4")),
    paste0("LRRTM",1:4)),
  neuromodulatory = c(
    paste0("DRD",1:5),
    paste0("HTR1",c("A","B","D","E","F")), paste0("HTR2",c("A","B","C")),
    paste0("HTR3",c("A","B","C","D","E")),
    paste0("HTR",4:7),
    paste0("CHRNA",1:10), paste0("CHRNB",1:4), paste0("CHRM",1:5),
    "NPY","PENK","TAC1","VIP"),
  glutamatergic = c(
    paste0("GRIA",1:4), "GRIN1", paste0("GRIN",2:3), paste0("GRIK",1:5),
    paste0("SLC1A",1:3), "GLUL","GLS","GLS2"),
  gabaergic = c(
    paste0("GABRA",1:6), paste0("GABRB",1:3), paste0("GABRG",1:3),
    "GABRD","GABBR1","GABBR2","SLC6A1","SLC6A11","GAD1","GAD2","ABAT"),
  cholinergic = c(
    paste0("CHRNA",1:10), paste0("CHRNB",1:4), paste0("CHRM",1:5),
    "SLC5A7","CHAT","ACHE"),
  serotonergic = c(
    paste0("HTR",1:7), "SLC6A4","TPH1","TPH2","MAOA","MAOB"),
  purinergic = c(
    paste0("P2RX",1:7),
    paste0("P2RY",1:14),
    paste0("ENTPD",1:3),
    "NT5E"),
  adrenergic = c("ADRA1A","ADRA1B","ADRA1D","ADRA2A","ADRA2B","ADRA2C",
                 paste0("ADRB",1:3),"SLC6A2","DBH","MAOA","MAOB","COMT")
)

category_df <- enframe(gene_sets, name = "category", value = "gene") %>%
  unnest(cols = "gene") %>%
  mutate(gene = toupper(gene))

category_df$unique <- isUnique(category_df$gene)
```

### Load DESeq2 Results

```{r load-deseq2}
# We will use the significant DE gene output files:
de_files <- c(
  "DESeq2_DIPG or DMG_paired_full.csv",
  "DESeq2_DIPG or DMG_vs_healthyNormal.csv",
  "DESeq2_EPN_paired_full.csv",
  "DESeq2_EPN_vs_healthyNormal.csv",
  "DESeq2_LGG_vs_healthyNormal.csv",
  "DESeq2_MB_paired_full.csv",
  "DESeq2_MB_vs_healthyNormal.csv"
)

de_files <- file.path(root_dir, "analyses", "human-rna-expression", "results", de_files)
```

```{r extract-tumor-group}
# A helper function identifies the tumor group from the filename:
extract_group <- function(filename) {
  if (grepl("DIPG|DMG", filename, ignore.case = TRUE)) return("DIPG_or_DMG")
  if (grepl("MB", filename, ignore.case = TRUE)) return("MB")
  if (grepl("EPN", filename, ignore.case = TRUE)) return("EPN")
  if (grepl("LGG", filename, ignore.case = TRUE)) return("LGG")
  return("Unknown")
}
```

### Generate Neuronal/Synaptic Gene Plots for Each Tumor Type

```{r generate-plot-per-histology}
# helper functions
reorder_within <- function(x, by, within, fun = mean, sep = "___") {
  new_x <- paste(x, within, sep = sep)
  stats <- tapply(by, new_x, fun)
  ordered <- names(sort(stats))
  factor(new_x, levels = ordered)
}

strip_within <- function(x, sep = "___") {
  gsub(paste0(sep, ".+$"), "", x)
}

all_categories <- unique(category_df$category)

# define your 3 groups of categories
cat_groups <- list(
  group1 = all_categories[1:4],
  group2 = all_categories[5:8],
  group3 = all_categories[9:11]
)

# -------------------------------
# 🔥 LOOP OVER DESeq2 FILES
# -------------------------------
for (f in de_files) {

  if (!file.exists(f)) next

  tumor_group <- extract_group(f)

  df <- read_csv(f) %>%
    mutate(gene = toupper(gene_symbol)) %>%
    select(gene, log2FoldChange, padj)

  # merge with neuronal categories
  merged_full <- category_df %>%
    mutate(category_wrapped = gsub("_", "\n", category)) %>%
    left_join(df, by = "gene") %>%
    filter(!is.na(padj), !is.na(log2FoldChange)) %>%
    mutate(
      de_group = case_when(
        padj < 0.05 & log2FoldChange > 0 ~ "Up",
        padj < 0.05 & log2FoldChange < 0 ~ "Down",
        TRUE ~ "Not_DE"
      ),
      size = -log10(padj),
      gene_ordered = reorder_within(gene, log2FoldChange, category)
    )

  if (nrow(merged_full) == 0) next

  # unified range for size scale
  size_min <- min(merged_full$size, na.rm = TRUE)
  size_max <- max(merged_full$size, na.rm = TRUE)

  # plotting function
  make_col_plot <- function(cat_list, show_legend = FALSE) {
    df_sub <- merged_full %>% filter(category %in% cat_list)

    ggplot(df_sub,
           aes(x = log2FoldChange, y = gene_ordered,
               size = size, fill = de_group)) +
      geom_point(shape = 21, color = "black", stroke = 0.3, alpha = 0.9) +
      facet_grid(category_wrapped ~ ., scales = "free_y",
                 space = "free_y", switch = "y") +
      scale_y_discrete(labels = function(x) strip_within(x)) +
      scale_fill_manual(values = c("Up"="green4","Down"="red3",
                                   "Not_DE"="whitesmoke"),
                        name="DE status") +
      scale_size_continuous(range=c(2,6),
                            limits=c(size_min,size_max),
                            name="-log10(padj)") +
      guides(fill = guide_legend(order=1, override.aes = list(shape = 21, size = 5, stroke = 0.6)),
             size = guide_legend(order=2)) +
      theme_Publication() +
      theme(
        strip.placement = "outside",
        strip.text.y.left = element_text(angle = 90, face = "bold", size = 14),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = ifelse(show_legend, "right", "none"),
        legend.title = element_text(size = 14, face = "bold"),
        legend.text  = element_text(size = 12)
      ) +
      labs(x="log2FoldChange", y=NULL)
  }

  # build 3-column layout
  p1 <- make_col_plot(cat_groups$group1, show_legend = FALSE)
  p2 <- make_col_plot(cat_groups$group2, show_legend = FALSE)
  p3 <- make_col_plot(cat_groups$group3, show_legend = TRUE)

  final_plot <- plot_grid(p1, p2, p3, ncol = 3,
                          rel_widths = c(1,1,1))

  # file naming
  base <- basename(f)
  comparison <- case_when(
    grepl("paired", base, ignore.case=TRUE) ~ "paired",
    grepl("healthyNormal", base, ignore.case=TRUE) ~ "healthyNormal",
    TRUE ~ "comparison"
  )

  out_file <- file.path(
    plot_dir,
    paste0("neuronal_", tumor_group, "_", comparison, ".pdf")
  )

  ggsave(out_file, final_plot, width = 18, height = 20)

  message("Saved: ", out_file)
}
```

### Session Info

```{r session-info}
sessionInfo()
```

