---
title: 'miRNA differential expression in CAR-T treated versus untreated mice'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2025"
---

This script performs differential expression analysis of human miRNAs in DIPG/DMG samples versus normal pons

Load packages

```{r load packages}
library(tidyverse)
library(DESeq2)
library(ggplot2)
library(EnhancedVolcano)
library(ggpubr)
```

Set directory paths

```{r dir paths}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "murine-mirna-differential-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

source(file.path(root_dir, "figures", "theme.R"))
```

Set file paths 

```{r file paths}
mirna_cts_file <- file.path(data_dir,
                          "30-963755216-miRNA_expression.csv")
```

Wrangle data. Remove Tube1 sample due to sparse counts as a results of low sequencing depth

```{r wrangle data}
mirna_cts <- read_csv(mirna_cts_file) %>%
  column_to_rownames("...1") %>% 
  dplyr::select(-`Day14-Tube1`)

names(mirna_cts) <- str_replace(names(mirna_cts), "RESUB-", "")

mirna_cts <- mirna_cts %>%
  filter(rowSums(across(everything(), ~ . != 0)) >= 5)
```

Create sample metaata df

```{r metadata}
metadata_df <- tibble(id = names(mirna_cts)) %>%
  dplyr::mutate(
    time = unlist(lapply(strsplit(id, "-"), function(x) x[[1]])),
    treatment = c(rep("B7H3", 2),
                  rep("STOP", 3), "untreated",
                  rep("B7H3", 3),
                  rep("STOP", 3), 
                  rep("untreated", 2),
                  rep("B7H3", 2), 
                  rep("STOP", 3), 
                  rep("untreated", 2))
  ) %>%
  dplyr::mutate(
    treatment = dplyr::recode(
      treatment,
      "B7H3" = "CAR",
      "STOP" = "Ctrl CAR",
      "untreated" = "Untreated"
    ),
    treatment = factor(treatment, levels = c("CAR", "Ctrl CAR", "Untreated")),
    time = factor(time, levels = c("Day14", "Day21", "Day28"))
  ) %>%
  as.data.frame()

write_tsv(metadata_df,
          file.path(results_dir,
                    "sample-metadata.tsv"))
```

Construct DESeqDataSet Object

```{r create dds}
dds_all <- DESeqDataSetFromMatrix(countData = mirna_cts, 
                              colData = metadata_df, 
                              design = ~ time + treatment)
```

Generate PCA plot from all samples, and by timepoint

```{r pca timepoints}
# transform raw counts
vsdata <- varianceStabilizingTransformation(dds_all)


p_all <- plotPCA(vsdata, intgroup = c("time", "treatment"), ntop = 250) +
  aes(color = treatment, shape = time) +
  scale_shape_manual(values = c("Day14" = 16, "Day21" = 17, "Day28" = 18)) +
  scale_color_manual(values = c("CAR" = "#E64B35FF",
                                "Ctrl CAR" = "#4DBBD5FF",
                                "Untreated" = "#00A087FF")) +
  theme_Publication() +
  labs(title = "All Samples", color = "Treatment", shape = "Time")

ggsave(file.path(plot_dir, glue::glue("mirna-pca-plot-all.pdf")),
       p_all, width = 5.5, height = 4.5)

# Get unique time points
time_points <- unique(colData(dds_all)$time)

p_list <- list()

# Create PCA plots for each time point
for (time in time_points) {
  
  # Subset data for the specific time point
  vsdata_subset <- vsdata[,colData(vsdata)$time == time]
  
  # Generate PCA plot
  p_list[[time]] <- plotPCA(vsdata_subset, intgroup = "treatment", ntop = 250) +
  aes(color = treatment) +
  scale_color_manual(values = c("CAR" = "#E64B35FF",
                                "Ctrl CAR" = "#4DBBD5FF",
                                "Untreated" = "#00A087FF")) +
  theme_Publication() +
  labs(title = time, color = "Treatment")
  
}

# merge PCA plots

p_timepoint <- ggarrange(p_list[["Day14"]],
               p_list[["Day21"]],
               p_list[["Day28"]],
               ncol = 1,
               align = "h")

# Save the plot
ggsave(file.path(plot_dir, glue::glue("mirna-pca-plot-by-timepoint.pdf")),
       p_timepoint, width = 5.5, height = 8.5)
```

Generate PCA plots by treatment

```{r pca treatments}
# Get unique treatments
treatments <- unique(colData(dds_all)$treatment)

p_list <- list()

# Create PCA plots for each time point
for (treatment in treatments) {
  
  # Subset data for the specific time point
  vsdata_subset <- vsdata[,colData(vsdata)$treatment == treatment]
  
  # Generate PCA plot
  p_list[[treatment]] <- plotPCA(vsdata_subset, intgroup = "time", ntop = 250) +
  ggtitle(glue::glue("PCA Plot - {treatment}")) +
  theme_Publication() +
  labs(title = treatment, color = "Time")

}

# merge treatment PCA plots

p_treatment <- ggarrange(p_list[["CAR"]],
               p_list[["Ctrl CAR"]],
               p_list[["Untreated"]],
               ncol = 1,
               align = "h")

# Save the plot
ggsave(file.path(plot_dir, glue::glue("mirna-pca-plot-by-treatment.pdf")),
       p_treatment, width = 5.5, height = 8.5)
```

For differential expression analyses, we will generate DESeqDataSet objects for each timepoint and perform the following comparisions:
- B7H3 versus untreated
- STOP versus untreated

```{r deseq2}
timepoints <- c("Day14", "Day21", "Day28")

# create empty results list
res <- list()

# loop through time points to perform differential expression analyses
for (timepoint in timepoints){

  # create DDS obejct
  dds <- DESeqDataSetFromMatrix(countData = mirna_cts[,metadata_df$time == timepoint], 
                                colData = metadata_df[metadata_df$time == timepoint,], 
                                design = ~ treatment)
  
  # run DESeq2
  dds <- DESeq(dds)
  
  # pull B7H3 versus untreated contrast and generate data frame
  res_B7H3_vs_Untreated <- results(dds, contrast = c("treatment", "CAR", "Untreated"))
  
  res_B7H3_vs_Untreated <- data.frame(res_B7H3_vs_Untreated) %>%
    rownames_to_column("mirna_id") %>%
    dplyr::arrange(pvalue) %>%
    dplyr::mutate(label = ifelse(dplyr::row_number() <= 5, mirna_id, ""))
  
  # generate volcano plot
  volc_plot <- EnhancedVolcano(
    res_B7H3_vs_Untreated,
    lab = res_B7H3_vs_Untreated$label,
    x = "log2FoldChange",
    y = "pvalue",
    title = glue::glue("CAR versus Untreated, {timepoint}"),
    subtitle = NULL,
    caption = NULL,
    pCutoff = max(res_B7H3_vs_Untreated$pvalue[res_B7H3_vs_Untreated$padj < 0.1], na.rm = TRUE),
    FCcutoff = 2,
    pointSize = 4,
    labSize = 4.5,
    typeConnectors = "closed",
    drawConnectors = TRUE,
    widthConnectors = 1,
    colConnectors = 'black',
    max.overlaps = 8
  )
  
  ggsave(file.path(
    plot_dir,
    glue::glue(
      "murine-mirna-deseq2-b7h3-versus-untreated-{timepoint}.pdf"
    )
  ),
  volc_plot,
  width = 7,
  height = 6)
  
  # pull STOP versus untreated contrast and generate data frame
  res_STOP_vs_untreated <- results(dds, 
                                  contrast = c("treatment", "Ctrl CAR", "Untreated"))
  
  res_STOP_vs_untreated <- data.frame(res_STOP_vs_untreated) %>%
    rownames_to_column("mirna_id")
  
  # join two contrast dfs by miRNA id, adding suffixes to column names to indicate comparisons
  res[[timepoint]] <- res_B7H3_vs_Untreated %>%
    left_join(res_STOP_vs_untreated,
              suffix = c("_b7h3_vs_untr",
                         "_stop_vs_untr"),
              by = "mirna_id") %>%
    # define B7H3-upregulated and -downregulated miRNAs as those DE in B7H3 versus untreated ONLY, and NOT DE in same dir in STOP versus untreated comparison 
    dplyr::mutate(status = case_when(
      padj_b7h3_vs_untr < 0.1 & log2FoldChange_b7h3_vs_untr > 0 & (padj_stop_vs_untr > 0.1 | log2FoldChange_stop_vs_untr < 0) ~ "B7H3 upregulated",
      padj_b7h3_vs_untr < 0.1 & log2FoldChange_b7h3_vs_untr < 0 & (padj_stop_vs_untr > 0.1  | log2FoldChange_stop_vs_untr > 0) ~ "B7H3 downregulated",
      TRUE ~ "Not DE"
    ))
  
}
```

Merge DE results by time point, and classify miRNAs based on differential expression patterns over time

```{r merge res}

# add "_Day28 suffix to Day28 df
names(res[["Day28"]])[-1] <- glue::glue("{names(res[['Day28']])[-1]}_Day28")

# join timepoint dfs
res_all <- res[["Day14"]] %>%
  left_join(res[["Day21"]],
            by = "mirna_id",
            suffix = c("_Day14", "_Day21")) %>%
    left_join(res[["Day28"]],
            by = "mirna_id") %>%
  # group miRNAs by differential expression patterns
  dplyr::mutate(expr_pattern = case_when(
    status_Day14 == "B7H3 upregulated" & status_Day21 == "Not DE" & status_Day28 == "Not DE" ~ "B7H3 up, Day 14",
    status_Day14 == "Not DE" & status_Day21 == "B7H3 upregulated" & status_Day28 == "Not DE" ~ "B7H3 up, Day 21",
    status_Day14 == "Not DE" & status_Day21 == "Not DE" & status_Day28 == "B7H3 upregulated" ~ "B7H3 up, Day 28",
    status_Day14 == "B7H3 downregulated" & status_Day21 == "Not DE" & status_Day28 == "Not DE" ~ "B7H3 down, Day 14",
    status_Day14 == "Not DE" & status_Day21 == "B7H3 downregulated" & status_Day28 == "Not DE" ~ "B7H3 down, Day 21",
    status_Day14 == "Not DE" & status_Day21 == "Not DE" & status_Day28 == "B7H3 downregulated" ~ "B7H3 down, Day 28",
    (status_Day14 == "B7H3 upregulated" & status_Day21 == "B7H3 upregulated") | (status_Day14 == "B7H3 upregulated" & status_Day28 == "B7H3 upregulated") | (status_Day21 == "B7H3 upregulated" & status_Day28 == "B7H3 upregulated") ~ "B7H3 up, multiple",
    (status_Day14 == "B7H3 downregulated" & status_Day21 == "B7H3 downregulated") | (status_Day14 == "B7H3 downregulated" & status_Day28 == "B7H3 downregulated") | (status_Day21 == "B7H3 downregulated" & status_Day28 == "B7H3 downregulated") ~ "B7H3 down, multiple",
    status_Day14 != "Not DE" | status_Day21 != "Not DE" | status_Day28 != "Not DE" ~ "Other DE",
    TRUE ~ "Not DE"
  ))
```

Print miRNAs DE at multiple time points

```{r print multi up}
res_all %>%
  dplyr::filter(expr_pattern == "B7H3 up, multiple") %>%
  dplyr::select(mirna_id, status_Day14,
                status_Day21, status_Day28)
```

```{r print multi down}
res_all %>%
  dplyr::filter(expr_pattern == "B7H3 down, multiple") %>%
  dplyr::select(mirna_id, status_Day14,
                status_Day21, status_Day28)
```
Print miRNAs with variable expression patterns across time points

```{r print variable de}
res_all %>%
  dplyr::filter(expr_pattern == "Other DE") %>%
  dplyr::select(mirna_id, status_Day14,
                status_Day21, status_Day28)
```

Save output

```{r save output}
write_tsv(res_all,
          file.path(results_dir,
                    "mirna-differential-expression-deseq2-b7h3-stop-vs-untreated-by-timepoint.tsv"))
```

Print session info

```{r session info}
sessionInfo()
```
