---
title: 'GO enrichment analysis of DE miRNA targets'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2025"
---

This script performs gene ontology (GO) enrichment analysis on DE miRNA predicted targets

Load packages

```{r load packages}
library(tidyverse)
library(ggplot2)
library(org.Mm.eg.db)
library(topGO)
library(AnnotationDbi)
```

Set directory paths

```{r dir paths}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "murine-mirna-differential-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")
```

Set file paths

```{r file paths}
de_mirna_file <- file.path(results_dir,
                           "mirna-differential-expression-deseq2-b7h3-stop-vs-untreated-by-timepoint.tsv")

mirTarBase_SE_R_file <- file.path(root_dir, "analyses",
                                  "human-mirna-rna-interactions", "input", "miRTarBase_SE_R.csv")
mirTarBase_SE_W_file <- file.path(root_dir, "analyses",
                                  "human-mirna-rna-interactions", "input", "miRTarBase_SE_W.csv")
mirTarBase_SE_WR_file <- file.path(root_dir, "analyses",
                                  "human-mirna-rna-interactions", "input", "miRTarBase_SE_WR.csv")
mirTarBase_WE_Clip_file <- file.path(root_dir, "analyses",
                                  "human-mirna-rna-interactions", "input", "miRTarBase_WE_Clip.tsv.gz")
mirTarBase_WE_Other_file <- file.path(root_dir, "analyses",
                                  "human-mirna-rna-interactions", "input", "miRTarBase_WE_Other.csv")
```

Load data and filter miRTarBase data frame for mouse miRNAs

```{r wrangle data}
de_mirna_df <- read_tsv(de_mirna_file)

mirTarBase_df <- read_csv(mirTarBase_SE_R_file) %>%
  bind_rows(read_csv(mirTarBase_SE_W_file)) %>%
  bind_rows(read_csv(mirTarBase_SE_WR_file)) %>%
  bind_rows(read_csv(mirTarBase_WE_Other_file)) %>%
  bind_rows(read_tsv(mirTarBase_WE_Clip_file)) %>%
  dplyr::filter(grepl("mmu", miRNA))
```

write mouse miRNA target prediction data frame to output

```{r write mirTarBase}
write_tsv(mirTarBase_df,
          file.path(results_dir,
                    "mouse-mirna-target-predictions-miRTarBase.tsv.gz"))
```

Extract known DE miRNAs (at any time point) 

```{r get DE miRNAs}
multiple_de_mirnas <- de_mirna_df %>%
  dplyr::filter(grepl("mmu", mirna_id),
                grepl("multiple", expr_pattern)) %>%
  pull(mirna_id)

d14_top_de_mirnas <- de_mirna_df %>%
  dplyr::filter(grepl("mmu", mirna_id),
                status_Day14 != "Not DE") %>%
  arrange(pvalue_b7h3_vs_untr_Day14) %>%
  head(10) %>%
  pull(mirna_id)

d21_top_de_mirnas <- de_mirna_df %>%
  dplyr::filter(grepl("mmu", mirna_id),
                status_Day21 != "Not DE") %>%
  arrange(pvalue_b7h3_vs_untr_Day21) %>%
  head(10) %>%
  pull(mirna_id)


d28_top_de_mirnas <- de_mirna_df %>%
  dplyr::filter(grepl("mmu", mirna_id),
                status_Day28 != "Not DE") %>%
  arrange(pvalue_b7h3_vs_untr_Day28) %>%
  head(10) %>%
  pull(mirna_id)

# add miR-155-5p (DE at 21d and a known regulator of immune response)
top_de_mirnas <- unique(c(multiple_de_mirnas,
                          d14_top_de_mirnas,
                          d21_top_de_mirnas,
                          d28_top_de_mirnas,
                          "mmu-miR-155-5p"))
```

Loop through DE miRNA, extract targets, and perform GO enrichment analyses

```{r known target GO}
# create empty lists
known_mirna_target_list <- list()
known_mirna_target_go_list <- list()

# get list of all genes to use as background for enrichment
all_entrez_ids <- keys(org.Mm.eg.db, keytype = "ENTREZID")

# Map Entrez IDs to Gene Symbols
all_genes <- select(org.Mm.eg.db, keys = all_entrez_ids, 
                    columns = "SYMBOL", keytype = "ENTREZID")

for (known_mirna in top_de_mirnas){
  
  # pull target genes from miRTarBase df
  known_mirna_target_list[[known_mirna]] <- mirTarBase_df %>%
    dplyr::filter(miRNA == known_mirna) %>%
    distinct(`Target Gene`) %>%
    pull(`Target Gene`)
  
  if (length(known_mirna_target_list[[known_mirna]] >= 20)){

    # define all genes as target (1) or not target (0) for enrichment
    target_gene_df <- all_genes %>%
    dplyr::mutate(target_gene = case_when(
      SYMBOL %in% known_mirna_target_list[[known_mirna]] ~ 1,
      TRUE ~ 0
    ))
  
    target_genes <- target_gene_df$target_gene
    names(target_genes) <- target_gene_df$SYMBOL
  
    # Create topGO object, using biological process (BP) ontology terms
    GO_data <- new("topGOdata", ontology = "BP", allGenes = target_genes, 
                                                     geneSel = function(x) x == 1, 
                                                     annot = annFUN.org, mapping = "org.Mm.eg.db", ID = "symbol")
    
    # Run Fisher test
    result <- runTest(GO_data, algorithm = "classic", statistic = "fisher")
    
    # write results to go analysis list
   known_mirna_target_go_list[[known_mirna]] <- GenTable(GO_data, classicFisher = result, topNodes = length(usedGO(GO_data)),
                                                         numChar = 100000)
   
   # filter for signficant results
   known_mirna_target_go_list[[known_mirna]] <- known_mirna_target_go_list[[known_mirna]] %>%
     dplyr::mutate(fdr = p.adjust(classicFisher, method = "BH")) %>%
     dplyr::filter(fdr < 0.05)
   
   # Add list of genes associated with each sig GO term to df

    genes_per_term <- genesInTerm(GO_data)
    
    genes_per_term_df <- enframe(genes_per_term, name = "GO.ID", value = "Genes") %>%
      mutate(Genes = sapply(Genes, function(x) paste(x, collapse = ", ")))
    
    known_mirna_target_go_list[[known_mirna]] <- known_mirna_target_go_list[[known_mirna]] %>%
      left_join(genes_per_term_df)
   
   if (nrow(known_mirna_target_go_list[[known_mirna]]) > 0){
   
     write_tsv(known_mirna_target_go_list[[known_mirna]],
               file.path(results_dir,
                         glue::glue("{known_mirna}-target-go-enrichment.tsv")))
     
   }
   
  }
  
}
```

Print session info

```{r session info}
sessionInfo()
```