---
title: "Differential RNA Expression Analysis"
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Bicna Song
date: "`r Sys.Date()`"
---

This script performs differential expression analysis of RNA-seq data using DESeq2.

### Setup

```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(DESeq2)
library(tidyverse)
library(rprojroot)
library(readr)
library(EnhancedVolcano)
library(ashr)
library(stringr)
library(ggpubr)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "murine-rna-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# File path for counts
gene_cts_file <- file.path(data_dir, "mouse-gene-counts-rsem-expected_count-collapsed.all.rds")

# File path for metadata
metadata_file <- file.path(analysis_dir, "input", "Haydar_Mouse_RNA_miRNA_manifest_IDs_assigned.tsv")
```

### Data Wrangling

```{r get-metadata}
metadata_df <- read_delim(metadata_file) %>%
  filter(experimental_strategy == "RNA-Seq") %>%
  select(external_sample_id, Day, Treatment, Bioassay_ID) %>%
  distinct(external_sample_id, .keep_all = TRUE) %>%
  rename(time = Day) %>%
  rename(treatment = Treatment) %>%
  mutate(
    time = paste0("Day", time),
    treatment = str_remove(treatment, "\\s*CAR\\b"),
    treatment = str_squish(treatment),
    treatment = if_else(treatment == "B7H3 STOP", "STOP", treatment)
  )
```

```{r load-and-filter-counts}
# load & filter genes with at least two non-zero samples
gene_cts <- readRDS(gene_cts_file) %>%
  filter(rowSums(across(everything(), ~ . != 0)) >= 5)

# subset to samples in metadata
valid_ids <- metadata_df$Bioassay_ID
gene_cts <- round(gene_cts[, intersect(colnames(gene_cts), valid_ids), drop = FALSE])
```

```{r align-metadata-to-counts}
metadata_df <- metadata_df %>%
  arrange(match(Bioassay_ID, names(gene_cts)))
```

### Construct DESeqDataSet Object

```{r message=FALSE, warning=FALSE}
dds_all <- DESeqDataSetFromMatrix(countData = gene_cts,
                                  colData = metadata_df,
                                  design = ~ time + treatment)
```

### Generate PCA plot from all samples

```{r}
# transform raw counts
vsdata <- varianceStabilizingTransformation(dds_all)
```

```{r}
# Build PCA data
pca_data <- plotPCA(vsdata, intgroup = c("time","treatment"), returnData = TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))

# add external_sample_id for labeling
pca_data$Bioassay_ID <- rownames(pca_data)
pca_data <- dplyr::left_join(
  pca_data,
  metadata_df[, c("Bioassay_ID","external_sample_id")],
  by = "Bioassay_ID"
)

jx <- 0.05 * diff(range(pca_data$PC1))
jy <- 0.05 * diff(range(pca_data$PC2))

cols_treat <- c(B7H3 = "#E64B35FF", STOP = "#4DBBD5FF", Untreated = "#00A087FF")

p_no_label <- ggplot(pca_data, aes(PC1, PC2, color = treatment, shape = time)) +
  geom_point(size = 3, alpha = 0.9,
             position = position_jitter(width = jx, height = jy)) +
  scale_shape_manual(values = c("Day14" = 16, "Day21" = 17)) +
  scale_color_manual(values = cols_treat) +
  labs(title = "All Samples",
       color = "Treatment", shape = "Time Point",
       x = paste0("PC1: ", percentVar[1], "% variance"),
       y = paste0("PC2: ", percentVar[2], "% variance")) +
  theme_Publication()

ggsave(file.path(plot_dir, "rna-pca-plot-all.pdf"),
       p_no_label, width = 5, height = 4)

# Labeled version
p_labeled <- p_no_label +
  ggrepel::geom_text_repel(aes(label = external_sample_id),
                           size = 3, max.overlaps = 20)
print(p_labeled)
```

### Generate PCA plot by timepoint

```{r}
# Get unique time points
time_points <- sort(unique(colData(dds_all)$time))

p_list <- list()

# Create PCA plots for each time point
for (time in time_points) {
  
  # Subset data for the specific time point
  vsdata_subset <- vsdata[,colData(vsdata)$time == time]
  metadata_subset <- metadata_df %>% dplyr::filter(time == time)
  
  pca_data <- plotPCA(vsdata_subset, intgroup = "treatment", returnData = TRUE)
  percentVar <- round(100 * attr(pca_data, "percentVar"))
  pca_data$Bioassay_ID <- rownames(pca_data)
  
  pca_data <- pca_data %>%
    dplyr::left_join(metadata_subset %>% dplyr::select(Bioassay_ID, external_sample_id),
                     by = "Bioassay_ID")
  
  p_no_label <- ggplot(pca_data, aes(PC1, PC2, color = treatment)) +
    geom_point(size = 3, alpha = 0.9,
               position = position_jitter(width = 0.2, height = 0.2)) +
    scale_color_manual(values = cols_treat) +
    labs(
      title = time, color = "Treatment",
      x = paste0("PC1: ", percentVar[1], "% variance"),
      y = paste0("PC2: ", percentVar[2], "% variance")
    ) +
    theme_Publication()
  
  p_labeled <- p_no_label +
  ggrepel::geom_text_repel(aes(label = external_sample_id),
                           size = 3, max.overlaps = 20)
  
  print(p_labeled)
  
  p_list[[time]] <- p_no_label
}
```

```{r}
# merge PCA plots
p_timepoint <- ggarrange(p_list[["Day14"]],
               p_list[["Day21"]],
               ncol = 1,
               align = "h")

# Save the plot
ggsave(file.path(plot_dir, glue::glue("rna-pca-plot-by-timepoint.pdf")),
       p_timepoint, width = 5, height = 6)
```

### Generate PCA plots by treatment

```{r}
treatments <- unique(colData(dds_all)$treatment)

p_list <- list()

for (treatment in treatments) {
  
  vsdata_subset <- vsdata[,colData(vsdata)$treatment == treatment]
  metadata_subset <- metadata_df %>% dplyr::filter(treatment == treatment)
  
  pca_data <- plotPCA(vsdata_subset, intgroup = "time", returnData = TRUE)
  percentVar <- round(100 * attr(pca_data, "percentVar"))
  pca_data$Bioassay_ID <- rownames(pca_data)
  
  pca_data <- pca_data %>%
    dplyr::left_join(metadata_subset %>% dplyr::select(Bioassay_ID, external_sample_id),
                     by = "Bioassay_ID")
  
  p_no_label <- ggplot(pca_data, aes(PC1, PC2, color = time)) +
    geom_point(size = 3, alpha = 0.9,
               position = position_jitter(width = 0.2, height = 0.2)) +
    scale_color_manual(values = cols_treat) +
    labs(
      title = treatment, color = "Time Point",
      x = paste0("PC1: ", percentVar[1], "% variance"),
      y = paste0("PC2: ", percentVar[2], "% variance")
    ) +
    theme_Publication()
  
  p_labeled <- p_no_label +
  ggrepel::geom_text_repel(aes(label = external_sample_id),
                           size = 3, max.overlaps = 20)
  
  print(p_labeled)
  
  p_list[[treatment]] <- p_no_label
}
```

```{r}
# merge PCA plots
p_timepoint <- ggarrange(p_list[["B7H3"]],
               p_list[["STOP"]],
               p_list[["Untreated"]],
               ncol = 1,
               align = "h")

# Save the plot
ggsave(file.path(plot_dir, glue::glue("rna-pca-plot-by-treatment.pdf")),
       p_timepoint, width = 5, height = 6)
```

### Run differential expression analyses

We will generate DESeqDataSet objects for each timepoint and perform the following comparisions:

- B7H3 versus untreated
- STOP versus untreated

```{r}
# create empty results list
res <- list()

genes_of_interest <- c("Cd8a", "Cd8b1", "Gzma", "Gzmb", "Prf1", "Ifng")

for (time in time_points){
  
  dds <- DESeqDataSetFromMatrix(countData = gene_cts[, metadata_df$time == time],
                                colData = metadata_df[metadata_df$time == time, ],
                                design = ~ treatment)
  
  dds <- DESeq(dds)
  
  # pull B7H3 versus untreated contrast and generate data frame
  res_B7H3_vs_Untreated <- results(dds, 
                                     contrast = c("treatment", "B7H3", "Untreated"))
  
  res_B7H3_vs_Untreated <- data.frame(res_B7H3_vs_Untreated) %>%
    rownames_to_column("gene_symbol")
  
  label_genes <- intersect(genes_of_interest, res_B7H3_vs_Untreated$gene_symbol)

  # generate volcano plot
  volc_plot <- EnhancedVolcano(res_B7H3_vs_Untreated,
                           lab = res_B7H3_vs_Untreated$gene_symbol, 
                           x = "log2FoldChange",
                           y = "pvalue",
                           title = glue::glue("B7H3 CAR-T versus Untreated, {time}"),
                           subtitle = NULL,
                           caption = NULL,
                           pCutoff = max(res_B7H3_vs_Untreated$pvalue[res_B7H3_vs_Untreated$padj < 0.05], na.rm = TRUE),
                           FCcutoff = 2,
                           pointSize = 4,
                           labSize = 4,
                           selectLab   = label_genes,
                           typeConnectors = "closed",
                           drawConnectors = TRUE,
                           widthConnectors = 1,
                           colConnectors = 'black',
                           max.overlaps = Inf)

  ggsave(file.path(plot_dir,
                glue::glue("murine-rna-deseq2-b7h3-versus-untreated-{time}.pdf")),
         volc_plot, width = 8, height = 6)
  
  
  # pull STOP versus untreated contrast and generate data frame
  res_STOP_vs_untreated <- results(dds, 
                                  contrast = c("treatment", "STOP", "Untreated"))
  
  res_STOP_vs_untreated <- data.frame(res_STOP_vs_untreated) %>%
    rownames_to_column("gene_symbol")
  
  # join two contrast dfs by gene_symbol, adding suffixes to column names to indicate comparisons
  res[[time]] <- res_B7H3_vs_Untreated %>%
    left_join(res_STOP_vs_untreated,
              suffix = c("_b7h3_vs_untr",
                         "_stop_vs_untr"),
              by = "gene_symbol") %>%
    # define B7H3-upregulated and -downregulated genes as those DE in B7H3 versus untreated ONLY, and NOT DE in STOP versus untreated comparison 
    dplyr::mutate(status = case_when(
      padj_b7h3_vs_untr < 0.05 & log2FoldChange_b7h3_vs_untr > 0 & padj_stop_vs_untr > 0.05 ~ "B7H3 upregulated",
      padj_b7h3_vs_untr < 0.05 & log2FoldChange_b7h3_vs_untr < 0 & padj_stop_vs_untr > 0.05 ~ "B7H3 downregulated",
      TRUE ~ "Not DE"
    ))
}
```

Merge DE results by time point, and classify genes based on differential expression patterns over time

```{r}
# join timepoint dfs
res_all <- res[["Day14"]] %>%
  left_join(res[["Day21"]],
            by = "gene_symbol",
            suffix = c("_Day14", "_Day21")) %>%
  # group genes by differential expression patterns
  dplyr::mutate(expr_pattern = case_when(
    status_Day14 == "B7H3 upregulated" & status_Day21 == "Not DE" ~ "B7H3 up, Day 14",
    status_Day14 == "Not DE" & status_Day21 == "B7H3 upregulated" ~ "B7H3 up, Day 21",
    status_Day14 == "B7H3 downregulated" & status_Day21 == "Not DE" ~ "B7H3 down, Day 14",
    status_Day14 == "Not DE" & status_Day21 == "B7H3 downregulated" ~ "B7H3 down, Day 21",
    status_Day14 == "B7H3 upregulated" & status_Day21 == "B7H3 upregulated" ~ "B7H3 up, multiple",
    status_Day14 == "B7H3 downregulated" & status_Day21 == "B7H3 downregulated" ~ "B7H3 down, multiple",
    status_Day14 != "Not DE" | status_Day21 != "Not DE" ~ "Other DE",
    TRUE ~ "Not DE"
  ))
```

Print DEGs at multiple time points

```{r}
res_all %>%
  dplyr::filter(expr_pattern == "B7H3 up, multiple") %>%
  dplyr::select(gene_symbol, status_Day14, status_Day21)
```

```{r}
res_all %>%
  dplyr::filter(expr_pattern == "B7H3 down, multiple") %>%
  dplyr::select(gene_symbol, status_Day14, status_Day21)
```

Print DEGs with variable expression patterns across time points

```{r}
res_all %>%
  dplyr::filter(expr_pattern == "Other DE") %>%
  dplyr::select(gene_symbol, status_Day14, status_Day21)
```

Save output

```{r}
write_tsv(res_all,
          file.path(results_dir,
                    "rna-differential-expression-deseq2-b7h3-stop-vs-untreated-by-timepoint.tsv"))
```

### Session Info

```{r session-info}
sessionInfo()
```
