---
title: 'B7H3 DE miRNA - immune target correlation analysis'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett, Bicna Song
date: "`r Sys.Date()`"
---

This script performs the following:  

1. Identifies B7H3 upregulated genes associated with enriched, CAR T therapy response GO terms
2. Identifies DE miRNAs that are predicted to target genes above, and that are downregualted in B7H3
3. Calculates correlation coefficients between miRNA-target pairs identified above
3. Plots most significant correlations

Load packages

```{r load packages}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(ggsci)
```

Set directory paths

```{r dir paths}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "murine-rna-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

source(file.path(root_dir, "figures", "theme.R"))
```

Set file paths

```{r}
de_mirna_file <- file.path(root_dir, "analyses",
                           "murine-mirna-differential-expression",
                           "results",
                           "mirna-differential-expression-deseq2-b7h3-stop-vs-untreated-by-timepoint.tsv")

de_rna_file <- file.path(root_dir, "analyses",
                           "murine-rna-expression",
                           "results",
                           "rna-differential-expression-deseq2-b7h3-stop-vs-untreated-by-timepoint.tsv")

mirna_target_file <- file.path(root_dir, "analyses", 
                               "murine-novel-mirna-target-prediction",
                               "results",
                               "mouse_miranda_output_parsed_anno.csv")

d14_up_go_file <- file.path(results_dir, 
                            "b7h3_up_d14-enriched-go-terms.tsv")

d21_up_go_file <- file.path(results_dir, 
                            "b7h3_up_d21-enriched-go-terms.tsv")

mirna_tpm_file <- file.path(root_dir, "analyses",
                            "murine-mirna-clustering",
                            "results",
                            "murine-mirna-tpm.rds")

rna_tpm_file <- file.path(data_dir, 
                          "mouse-gene-expression-rsem-tpm-collapsed.rds")

meta_file <- file.path(root_dir, "analyses",
                       "murine-rna-expression",
                       "input",
                       "Haydar_Mouse_RNA_miRNA_manifest_IDs_assigned.tsv")
```

Wrangle data

```{r}
# DE miRNA and gene results
de_mirna_df <- read_tsv(de_mirna_file) %>%
  dplyr::filter(expr_pattern != "Not DE") 

de_rna_df <- read_tsv(de_rna_file)

# miRNA target df
mirna_target_df <- read_csv(mirna_target_file) %>%
  # filter for high confidence targets
  filter(Max_Score >= 145, Max_Energy <= -20,
        GeneSymbol %in% de_rna_df$gene_symbol)

# GO enrichment results
d14_up_go_res <- read_tsv(d14_up_go_file)

d21_up_go_res <- read_tsv(d21_up_go_file)
```

Filter GO enrichment results for immune response terms, and extract DEGs per term

```{r}
keywords <- "immunity|granzyme|T cell|immune|leukocyte|lymphocyte|myeloid|monocyte|chemokine|cytokine|inflammatory|cell killing|chemotaxis|T-helper|toll-like|inflammasome|natural killer|transforming growth factor beta|MHC|antigen|IP-10|interleukin|interferon|thymocyte|macrophage|B cell"

d14_up_immune_go <- d14_up_go_res %>%
  dplyr::filter(grepl(keywords, Term))

d14_up_immune_genes <- unique(unlist(strsplit(d14_up_immune_go$DEGs, ", ")))

d21_up_immune_go <- d21_up_go_res %>%
  dplyr::filter(grepl(keywords, Term))

d21_up_immune_genes <- unique(unlist(strsplit(d21_up_immune_go$DEGs, ", ")))
```

Identify downregulated miRNAs predictd to target any of the d14 or d21 upregulated genes

```{r}
down_mirnas <- de_mirna_df %>%
  dplyr::filter(grepl("down, Day 14|down, Day 21", expr_pattern)) %>%
  pull(mirna_id)


de_target_df <- mirna_target_df %>% 
  dplyr::filter(GeneSymbol %in% c(d14_up_immune_genes,
                                  d21_up_immune_genes),
                Seq1 %in% down_mirnas) %>%
  distinct(Seq1, GeneSymbol, .keep_all = TRUE)
```

Load expression matrices

```{r}
mirna_tpm_mat <- readRDS(mirna_tpm_file)

rna_tpm_mat <- readRDS(rna_tpm_file) 
names(rna_tpm_mat) <- str_remove(names(rna_tpm_mat), "_[^_]+$")
```

Load sample metadata to match RNA IDs to miRNA IDs for correlation analyses

```{r}
meta_df <- read_tsv(meta_file)

match_df <- meta_df %>% 
  # create ID columns for matching
  dplyr::mutate(id = glue::glue("{Day}_{Treatment}_{external_aliquot_id}")) %>%
  # pull miRNA data
  dplyr::filter(experimental_strategy == "miRNA-Seq",
                !grepl("28", external_sample_id),
                external_sample_id != "Day14-Tube1") %>%
  distinct(id, external_sample_id) %>%
  dplyr::rename(mirna_id = external_sample_id) %>%
  # Join RNA sample data
  left_join(meta_df %>%
              dplyr::filter(experimental_strategy == "RNA-Seq",
                !grepl("28", external_sample_id)) %>%
              dplyr::mutate(id = glue::glue("{Day}_{Treatment}_{external_aliquot_id}")) %>%
                distinct(id, Bioassay_ID)) %>%
  dplyr::rename(rna_id = Bioassay_ID) %>%
  # filter for samples without matched miRNA & RNA-seq
  dplyr::filter(!is.na(mirna_id) & !is.na(rna_id)) %>%
  # define trt & timepoint columns
  dplyr::mutate(Timepoint = unlist(lapply(strsplit(mirna_id, "-"), function(x) x[[1]])),
                Treatment = case_when(
                  grepl("STOP", id) ~ "Ctrl CAR",
                  grepl("B7H3", id) ~ "CAR",
                  grepl("Untreated", id) ~ "Untreated"
                ))

# order miRNA and RNA tpm columns to match for correlation analyses
mirna_tpm_mat <- mirna_tpm_mat %>%
  dplyr::select(any_of(match_df$mirna_id))

rna_tpm_mat <- rna_tpm_mat %>%
  dplyr::select(any_of(match_df$rna_id))
```

Run correlation analyses

```{r}
de_target_df <- de_target_df %>%
  dplyr::mutate(pearson_r = NA_integer_,
                pearson_p = NA_integer_)

# Loop through rows of target df
for (i in 1:nrow(de_target_df)){
  
  mirna <- de_target_df$Seq1[i]
  target <- de_target_df$GeneSymbol[i]
  
  #pull tpms
  mirna_tpms <- log2(unlist(mirna_tpm_mat[mirna,] + 1))
  target_tpms <- log2(unlist(rna_tpm_mat[target,] + 1))
  
  # calculate pearson correlation coefficients and p-values
  de_target_df$pearson_r[i] <- cor.test(mirna_tpms,
                                        target_tpms,
                                        method = "pearson")$estimate
  de_target_df$pearson_p[i] <- cor.test(mirna_tpms,
                                      target_tpms,
                                        method = "pearson")$p.value
  
  # plot most significant correlations
  if (de_target_df$pearson_p[i] < 6e-4){
    
    plot_df <- match_df %>%
      dplyr::mutate(mirna_tpm = mirna_tpms,
                    target_tpm = target_tpms)
    
    plot_df %>%
      
      ggplot(aes(x = mirna_tpm, y = target_tpm)) +
      geom_point(aes(color = Treatment,
                     shape = Timepoint),
                 show.legend = TRUE) +
      stat_smooth(method = "lm", 
                  formula = y ~ x, 
                  geom = "smooth", 
                  colour = "red",
                  fill = "pink",
                  linetype="dashed") +
      labs(x = glue::glue("{mirna}\nlog2-TPM"),
           y = glue::glue("{target}\nlog2-TPM"),
           color = "Treatment",
           shape = "Timepoint") + 
      stat_cor(method = "pearson",
               label.x.npc = "center", 
               label.y.npc = "top", size = 2.25,
               p.digits = 3) +
      scale_color_npg() +
      theme_Publication()
    
    ggsave(file.path(plot_dir,
                     glue::glue("{mirna}-{target}-tpm-dotplot.pdf")),
           width = 4.5, height = 3)
    
  }
  
}

de_target_df <- de_target_df %>%
  dplyr::mutate(pearson_fdr = p.adjust(pearson_p, method = "BH"))
```

Print session info
```{r}
sessionInfo()
```

