---
title: "Merge miRNA count data"
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Bicna Song
date: "2025-05-09"
---

This script merges two batches of miRNA expression count data, standardizes novel miRNA IDs using mature and precurosr sequence information, and saves the merged output for downstream analysis.

### Load required packages

```{r load packages}
library(readr)
library(readxl)
library(tidyverse)
library(rprojroot)
```

### Set directory paths

```{r dir paths}
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-mirna-expression")
results_dir <- file.path(analysis_dir, "results")
```

### Set file paths 

```{r file paths}
batch1_counts_file <- file.path(data_dir, "30-931106737-miRNA_expression.csv")
batch2_counts_file <- file.path(data_dir, "30-1075661268-miRNA_expression.xls")
batch1_novel_mirna_file <- file.path(data_dir, "all_novel_miRNA.xls")
batch2_novel_mirna_file <- file.path(data_dir, "30-1075661268-all_novel_miRNA.xls")
```

### Load count matrices

```{r load count matrices}
batch1_counts <- read_csv(batch1_counts_file) %>%
  rename(miRNA = "...1") %>%
  column_to_rownames(var = "miRNA")

batch2_counts <- read_delim(batch2_counts_file, delim = "\t", guess_max = 10000) %>%
  rename(miRNA = "...1") %>%
  column_to_rownames(var = "miRNA")
```


### Load and process novel miRNA annotations

```{r load novel miRNA sequence}
read_novel_mirna <- function(file) {
  read_delim(file, delim = "\t", guess_max = 10000,
             col_names = c("id", "mature_seq", "star_seq", "precursor_seq", "precursor_coordinate")) %>%
    mutate(unique_id = paste0(mature_seq, "_", precursor_coordinate))
}

batch1_novel_mirna_seq <- read_novel_mirna(batch1_novel_mirna_file)
batch2_novel_mirna_seq <- read_novel_mirna(batch2_novel_mirna_file)
```

### Replace novel miRNA IDs with unique sequences

```{r replace novel miRNA ids}
replace_novel_ids <- function(counts, novel_df) {
  novel_lookup <- setNames(novel_df$unique_id, novel_df$id)
  current_names <- rownames(counts)
  is_novel <- grepl("^NovelmiRNA", current_names)
  
  # Warn if any novel miRNAs aren't found in the annotation
  unmapped <- setdiff(current_names[is_novel], names(novel_lookup))
  if (length(unmapped) > 0) {
    warning("Unmapped NovelmiRNA IDs: ", paste(unmapped, collapse = ", "))
  }
  
  current_names[is_novel] <- novel_lookup[current_names[is_novel]]
  rownames(counts) <- current_names
  return(counts)
}

batch1_counts <- replace_novel_ids(batch1_counts, batch1_novel_mirna_seq)
batch2_counts <- replace_novel_ids(batch2_counts, batch2_novel_mirna_seq)
```

### Merge counts matrices

```{r merge counts}
# Get union of all miRNA IDs
all_miRNAs <- union(rownames(batch1_counts), rownames(batch2_counts))

# Create full matrices with all miRNAs 
batch1_full <- matrix(0, nrow = length(all_miRNAs), ncol = ncol(batch1_counts),
                      dimnames = list(all_miRNAs, colnames(batch1_counts)))
batch2_full <- matrix(0, nrow = length(all_miRNAs), ncol = ncol(batch2_counts),
                      dimnames = list(all_miRNAs, colnames(batch2_counts)))

# Fill in actual counts
batch1_full[rownames(batch1_counts), ] <- as.matrix(batch1_counts)
batch2_full[rownames(batch2_counts), ] <- as.matrix(batch2_counts)

# Merge matrices
merged_counts <- as.data.frame(cbind(batch1_full, batch2_full))
```

### Save merged output

```{r write output}
# Save merged matrix with rownames as a column
merged_counts_out <- merged_counts %>%
  rownames_to_column(var = "miRNA")

write_tsv(merged_counts_out, file.path(results_dir, "merged-miRNA-expression.tsv"))
```

### Session info

```{r session info}
sessionInfo()
```

