---
title: "miRNA Expression PCA: Uncorrected vs Batch Corrected"
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Bicna Song
date: "2025-05-15"
---

This script compares PCA plots of miRNA expression data before and after batch correction.

### Load required packages

```{r load packages}
library(DESeq2)
library(tidyverse)
library(limma)
library(ggplot2)
library(rprojroot)
library(patchwork)
```

### Set directory paths

```{r dir paths}
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-mirna-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

source(file.path(root_dir, "figures", "theme.R"))
```

### Set file paths 

```{r file paths}
mirna_cts_file <- file.path(results_dir, "merged-miRNA-expression.tsv")
```

### Wrangle data

```{r wrangle data}
mirna_cts <- read_tsv(mirna_cts_file) %>%
  column_to_rownames("miRNA") %>%
  filter(rowSums(across(everything(), ~ . != 0)) >= 2)
```

### Load sample metadata df

```{r metadata}
metadata_df <- read_delim(file.path(data_dir, "miRNA-sample-metadata.txt")) %>%
  mutate(histology = if_else(is.na(histology) | histology == "N/A", "Normal", histology)) %>%
  filter(sample_id %in% names(mirna_cts)) %>%
  arrange(match(sample_id, names(mirna_cts))) %>%
  mutate(batch = if_else(row_number() <= 5, "batch1", "batch2"),
         condition = case_when(
    sample_type %in% c("Tumor", "Normal") ~ histology,
    sample_type == "Normal-Tumor Adjacent" ~ paste0(histology, "-adjacent Normal"),
    TRUE ~ NA_character_
  ))
```

### Construct DESeqDataSet Object

```{r create dds}
# 
dds <- DESeqDataSetFromMatrix(countData = mirna_cts, 
                              colData = metadata_df, 
                              design = ~ batch + condition)
```

### Variance Stabilizing Transformation (VST)

```{r vst-transform-1}
vsdata <- varianceStabilizingTransformation(dds, blind = FALSE)
```

### PCA Plot: Uncorrected Data

```{r pca-uncorrected-1}
pca_uncorrected <- plotPCA(vsdata, intgroup = c("batch", "condition"), returnData = TRUE)
percentVar <- round(100 * attr(pca_uncorrected, "percentVar"))

p1 <- ggplot(pca_uncorrected, aes(PC1, PC2, color = condition, shape = batch)) +
  geom_point(size = 3) +
  labs(title = "PCA (Uncorrected)",
       x = paste0("PC1: ", percentVar[1], "% variance"),
       y = paste0("PC2: ", percentVar[2], "% variance")) +
  theme_Publication()
```

### PCA Plot: Batch-Corrected Data

```{r pca-corrected}
# Apply batch correction to vst data
corrected_counts <- removeBatchEffect(assay(vsdata), batch = vsdata$batch)

# PCA
pca_corrected <- prcomp(t(corrected_counts))

# Prepare data
pca_df <- as.data.frame(pca_corrected$x)
pca_df$condition <- colData(vsdata)$condition
pca_df$batch <- colData(vsdata)$batch

percentVar_corrected <- round(100 * (pca_corrected$sdev^2 / sum(pca_corrected$sdev^2)))

p2 <- ggplot(pca_df, aes(PC1, PC2, color = condition, shape = batch)) +
  geom_point(size = 3) +
  labs(title = "PCA (Batch Corrected)",
       x = paste0("PC1: ", percentVar_corrected[1], "% variance"),
       y = paste0("PC2: ", percentVar_corrected[2], "% variance")) +
  theme_Publication()
```

### Save Plots

```{r save-plots}
ggsave(filename = file.path(plot_dir, "mirna-pca-uncorrected.pdf"), plot = p1, width = 10, height = 6)
ggsave(filename = file.path(plot_dir, "mirna-pca-corrected.pdf"), plot = p2, width = 10, height = 6)
```

# Run more stringent filtering of miRNA counts, retaining only those with non-zero counts in >50 samples

```{r}
mirna_cts_n50 <- read_tsv(mirna_cts_file) %>%
  column_to_rownames("miRNA") %>%
  filter(rowSums(across(everything(), ~ . != 0)) > 50)
```

### Create updated DESeqDataSet object

```{r create dds filtered}
# 
dds_filtered <- DESeqDataSetFromMatrix(countData = mirna_cts_n50, 
                              colData = metadata_df, 
                              design = ~ batch + condition)
```

### Variance Stabilizing Transformation (VST)

```{r vst-transform-2}
vsdata_filtered <- varianceStabilizingTransformation(dds_filtered, blind = FALSE)
```

### PCA Plot: Uncorrected, filtered Data

```{r pca-uncorrected-2}
pca_uncorrected_filtered <- plotPCA(vsdata_filtered, intgroup = c("batch", "condition"), returnData = TRUE)
percentVar_filtered <- round(100 * attr(pca_uncorrected_filtered, "percentVar"))

p3 <- ggplot(pca_uncorrected_filtered, aes(PC1, PC2, color = condition, shape = batch)) +
  geom_point(size = 3) +
  labs(title = "PCA (Uncorrected)",
       x = paste0("PC1: ", percentVar_filtered[1], "% variance"),
       y = paste0("PC2: ", percentVar_filtered[2], "% variance")) +
  theme_Publication()

ggsave(filename = file.path(plot_dir, "mirna-pca-uncorrected_filtered_n50.pdf"), plot = p3, width = 10, height = 6)
```


### Session info

```{r session info}
sessionInfo()
```