---
title: "Differential miRNA Expression Analysis"
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Bicna Song
date: "2025-05-23"
---

This script performs differential expression analysis of miRNA-seq data using DESeq2.

The goal is to identify differentially expressed miRNAs for each brain tumor histology by comparing:

- Tumor vs. Tumor-Adjacent Normal samples (e.g., DIPG vs DIPG-adjacent normal)
- Tumor vs. Disease-Free Normal brain samples (B1–B5)

### Setup
```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(DESeq2)
library(tidyverse)
library(rprojroot)
library(readr)
library(EnhancedVolcano)

#install.packages("ashr") # this will be added to Dockerfile later if needed.
library(ashr)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-mirna-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# File path for counts
mirna_cts_file <- file.path(results_dir, "merged-miRNA-expression.tsv")

# File path for histology
histology_file <- file.path(root_dir, "analyses", "histology-preprocessing", "results", "histologies.tsv")
```

### Data Wrangling

```{r data-wrangle}
mirna_cts <- read_tsv(mirna_cts_file) %>%
  column_to_rownames("miRNA") %>%
  filter(rowSums(across(everything(), ~ . != 0)) >= 2)
```

```{r rename-sample}
names(mirna_cts)[1] <- "4-1958-CC1"
names(mirna_cts)[2] <- "2-1578-CC1"
names(mirna_cts)[3] <- "5-1234-left-pons"
names(mirna_cts)[4] <- "1-1855-CC1"
names(mirna_cts)[5] <- "3-1890-CC1"
```

```{r remove-tumor-infiltrated-sample}
# Remove 5-1234-left-pons showing significant tumor infiltration
mirna_cts <- mirna_cts[, !(colnames(mirna_cts) == "5-1234-left-pons")]
```

```{r fix-tumor-normal-swap}
# Get current column names
current_colnames <- colnames(mirna_cts)

# Find the indices of the two samples
idx_T <- which(current_colnames == "2058-T")
idx_N <- which(current_colnames == "2058-N")

colnames(mirna_cts)[c(idx_T, idx_N)] <- colnames(mirna_cts)[c(idx_N, idx_T)]
```

```{r load-histology}
# Load and annotate histology file
histology_df <- read_delim(histology_file) %>%
  filter(experimental_strategy == "miRNA-Seq") %>%
  filter(external_sample_id %in% names(mirna_cts)) %>%
  arrange(match(external_sample_id, names(mirna_cts))) %>%
  # Fill in any missing histology with "Normal"
  mutate(histology = if_else(is.na(histology), "Normal", histology)) %>%
  mutate(
    batch             = if_else(row_number() <= 4, "batch1", "batch2"),
    condition         = case_when(
      sample_type %in% c("Tumor", "Normal") ~ histology,
      sample_type == "Normal-Tumor Adjacent" ~ paste0(histology, "-adjacent Normal"),
      TRUE ~ NA_character_
    ),
    histology_condition = case_when(
      sample_type %in% c("Tumor", "Normal") ~ histology,
      sample_type == "Normal-Tumor Adjacent" ~ "Normal-Tumor Adjacent",
      TRUE ~ NA_character_
    )
  )
```

### Perform Differential Expression for Each Histology

```{r run-de-by-histology, message=FALSE, warning=FALSE}
normal_ids <- c("B1", "B2", "B3", "B4", "B5")
histologies <- c("ATRT", "DIPG or DMG", "EPN", "HGG", "MB")

for (tumor in histologies) {
  adj_label <- paste0(tumor, "-adjacent Normal")

  # Subset metadata
  meta_adj <- histology_df %>% filter(condition %in% c(tumor, adj_label))
  meta_healthy <- histology_df %>% filter(external_sample_id %in% c(normal_ids, external_sample_id[condition == tumor])) %>% 
    filter(condition %in% c(tumor, "Normal"))

  # Select count data
  counts_adj <- mirna_cts[, meta_adj$external_sample_id]
  counts_healthy <- mirna_cts[, meta_healthy$external_sample_id]

  # Tumor vs Adjacent Normal
  design_adj <- if (tumor == "DIPG or DMG") ~ batch + condition else ~ condition
  dds_adj <- DESeqDataSetFromMatrix(countData = counts_adj, colData = meta_adj, design = design_adj)
  dds_adj$condition <- droplevels(dds_adj$condition)
  dds_adj <- DESeq(dds_adj)
  res_adj <- results(dds_adj, contrast = c("condition", tumor, adj_label))
  res_adj <- lfcShrink(dds_adj, contrast = c("condition", tumor, adj_label), res = res_adj, type = "ashr")
  write_csv(as.data.frame(res_adj) %>% rownames_to_column("miRNA"),
          file.path(results_dir, paste0("DESeq2_", tumor, "_vs_adjNormal.csv")))

  # Tumor vs Healthy Normal
  design_healthy <- if (tumor == "DIPG or DMG") ~ batch + condition else ~ condition
  dds_healthy <- DESeqDataSetFromMatrix(countData = counts_healthy, colData = meta_healthy, design = design_healthy)
  dds_healthy$condition <- droplevels(dds_healthy$condition)
  dds_healthy <- DESeq(dds_healthy)
  res_healthy <- results(dds_healthy, contrast = c("condition", tumor, "Normal"))
  res_healthy <- lfcShrink(dds_healthy, contrast = c("condition", tumor, "Normal"), res = res_healthy, type = "ashr")
  write_csv(as.data.frame(res_healthy) %>% rownames_to_column("miRNA"),
          file.path(results_dir, paste0("DESeq2_", tumor, "_vs_healthyNormal.csv")))
}
```

```{r run-de-lgg, message=FALSE, warning=FALSE}
# Handle LGG separately: only Tumor vs Healthy Normal
tumor <- "LGG"
meta_lgg <- histology_df %>%
  filter(external_sample_id %in% c(normal_ids, external_sample_id[condition == tumor])) %>%
  filter(condition %in% c(tumor, "Normal"))
counts_lgg <- mirna_cts[, meta_lgg$external_sample_id]

design_lgg <- ~ condition
dds_lgg <- DESeqDataSetFromMatrix(countData = counts_lgg, colData = meta_lgg, design = design_lgg)

dds_lgg$condition <- droplevels(dds_lgg$condition)
dds_lgg <- DESeq(dds_lgg)

res_lgg <- results(dds_lgg, contrast = c("condition", tumor, "Normal"))
res_lgg <- lfcShrink(dds_lgg, contrast = c("condition", tumor, "Normal"), res = res_lgg, type = "ashr")

write_csv(as.data.frame(res_lgg) %>% rownames_to_column("miRNA"),
          file.path(results_dir, paste0("DESeq2_", tumor, "_vs_healthyNormal.csv")))
```

```{r save-sig-de-res}
# Save Significant DE Results Separately
all_histologies <- c("ATRT", "DIPG or DMG", "EPN", "HGG", "MB", "LGG")

for (tumor in all_histologies) {
  comparison_types <- if (tumor == "LGG") "healthyNormal" else c("adjNormal", "healthyNormal")
  
  for (type in comparison_types) {
    file_path <- file.path(results_dir, paste0("DESeq2_", tumor, "_vs_", type, ".csv"))
    if (file.exists(file_path)) {
      res_df <- read_csv(file_path)
      sig_df <- res_df %>% filter(!is.na(padj) & padj < 0.05)
      write_csv(sig_df, file.path(results_dir, paste0("DESeq2_", tumor, "_vs_", type, "_sig.csv")))
    }
  }
}
```

### Summary Table: Number of Up/Down-regulated miRNAs

```{r summary-table}
summary_df <- tibble()

for (tumor in all_histologies) {
  comparison_types <- if (tumor == "LGG") "healthyNormal" else c("adjNormal", "healthyNormal")
  
  for (type in comparison_types) {
    file_path <- file.path(results_dir, paste0("DESeq2_", tumor, "_vs_", type, ".csv"))
    if (file.exists(file_path)) {
      res_df <- read_csv(file_path) %>%
        filter(!is.na(padj) & padj < 0.05)
      up <- sum(res_df$log2FoldChange > 0)
      down <- sum(res_df$log2FoldChange < 0)
      summary_df <- bind_rows(summary_df, 
        tibble(Comparison = paste0(tumor, " vs ", type), 
               Upregulated = up, 
               Downregulated = down))
    }
  }
}

summary_df

# Save to CSV
write_csv(summary_df, file.path(results_dir, "DE_summary_counts.csv"))
```

### Volcano Plots for DE Comparisons

```{r volcano-plots, message=FALSE, warning=FALSE}
for (tumor in all_histologies) {
  comparison_types <- if (tumor == "LGG") "healthyNormal" else c("adjNormal", "healthyNormal")
  
  for (type in comparison_types) {
    file_path <- file.path(results_dir, paste0("DESeq2_", tumor, "_vs_", type, ".csv"))
    if (file.exists(file_path)) {
      res_df <- read_csv(file_path)
      rownames(res_df) <- res_df$miRNA %||% rownames(res_df)
      p_cut <- max(res_df$pvalue[res_df$padj < 0.05], na.rm = TRUE)
      comp_label <- paste(tumor, "vs", gsub("Normal", "normal", type))
      
      plot <- EnhancedVolcano(res_df,
                              lab = rownames(res_df),
                              x = "log2FoldChange",
                              y = "pvalue",
                              title = comp_label,
                              subtitle = NULL,
                              caption = NULL,
                              pCutoff = p_cut,
                              FCcutoff = 2,
                              pointSize = 3.5,
                              labSize = 3,
                              typeConnectors = "closed",
                              drawConnectors = TRUE,
                              widthConnectors = 0.75,
                              colConnectors = 'grey30',
                              max.overlaps = 10)

      ggsave(file.path(plot_dir, paste0("volcano_", tumor, "_vs_", type, ".pdf")),
             plot = plot, width = 8, height = 6)
    }
  }
}
```

```{r pair-status}
# Summarize: Count number of Tumor and Adjacent Normal per sample_id and histology
paired_info <- histology_df %>%
  group_by(histology, sample_id) %>%
  summarize(
    tumor = any(sample_type == "Tumor"),
    adjacent_normal = any(sample_type == "Normal-Tumor Adjacent"),
    .groups = "drop"
  ) %>%
  mutate(pair_status = case_when(
    tumor & adjacent_normal ~ "Paired",
    TRUE ~ "Unpaired"
  ))

# Create table of paired/unpaired counts per histology
summary_table <- paired_info %>%
  group_by(histology, pair_status) %>%
  summarize(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = pair_status, values_from = n, values_fill = 0)

print(summary_table)
```

```{r run-paired-de-by-histology}
# Paired differential expression analysis
for (hist in all_histologies) {
  
  paired_patients <- paired_info %>%
    filter(histology == hist, pair_status == "Paired") %>%
    pull(sample_id)
  
  if (length(paired_patients) < 2) {
    message(paste("Skipping", hist, "due to insufficient paired patients"))
    next
  }
  
  paired_samples_metadata <- histology_df %>%
    filter(histology == hist, sample_id %in% paired_patients) %>%
    filter(sample_type %in% c("Tumor", "Normal-Tumor Adjacent"))
  
  paired_external_sample_ids <- paired_samples_metadata$external_sample_id
  
  mirna_cts_paired <- mirna_cts[, paired_external_sample_ids]
  
  # Confirm external_sample_id matches column names
  stopifnot(identical(paired_samples_metadata$external_sample_id, colnames(mirna_cts_paired)))
  
  # Create DESeq2 object
  dds <- DESeqDataSetFromMatrix(
    countData = mirna_cts_paired,
    colData = paired_samples_metadata,
    design = ~ sample_id + sample_type
  )
  
  #dds$sample_type <- factor(dds$sample_type)
  dds <- DESeq(dds)
  
  # Run results: Tumor vs Normal-Tumor Adjacent
  res <- results(dds, contrast = c("sample_type", "Tumor", "Normal-Tumor Adjacent"))
  res_adj <- lfcShrink(dds, contrast = c("sample_type", "Tumor", "Normal-Tumor Adjacent"), res = res, type = "ashr")
  
  # Save full and significant results
  res_df <- as.data.frame(res_adj) %>% 
    rownames_to_column("miRNA")
  
  write_csv(res_df, file.path(results_dir, paste0("DESeq2_", hist, "_paired_full.csv")))
  
  res_sig <- res_df %>%
    filter(!is.na(padj) & padj < 0.05)
  
  write_csv(res_sig, file.path(results_dir, paste0("DESeq2_", hist, "_paired_sig.csv")))
  
  message(paste("Completed DE analysis for:", hist))
}

```

### Summary Table: Number of Up/Down-regulated miRNAs for paired analysis

```{r}
summary_paired_df <- tibble()

for (hist in all_histologies) {
  file_path <- file.path(results_dir, paste0("DESeq2_", hist, "_paired_full.csv"))
  
  if (file.exists(file_path)) {
      res_df <- read_csv(file_path) %>%
        filter(!is.na(padj) & padj < 0.05)
      up <- sum(res_df$log2FoldChange > 0)
      down <- sum(res_df$log2FoldChange < 0)
      summary_paired_df <- bind_rows(summary_paired_df, 
        tibble(Comparison = paste0(hist, " vs ", "adjNormal"), 
               Upregulated = up, 
               Downregulated = down))
  } else {
    message(paste("Result file for", hist, "not found."))
  }
}

# View the summary table
summary_paired_df

# Save to CSV
write_csv(summary_paired_df, file.path(results_dir, "DE_summary_counts_paired.csv"))
```

### Volcano Plots for paired DE Comparisons

```{r}
for (tumor in all_histologies) {
  file_path <- file.path(results_dir, paste0("DESeq2_", tumor, "_paired_full.csv"))
  
  if (file.exists(file_path)) {
    res_df <- read_csv(file_path)
    rownames(res_df) <- res_df$miRNA %||% rownames(res_df)
    p_cut <- max(res_df$pvalue[res_df$padj < 0.05], na.rm = TRUE)
    comp_label <- paste(tumor, "Paired Analysis")
    plot <- EnhancedVolcano(res_df,
                             lab = rownames(res_df),
                             x = "log2FoldChange",
                             y = "pvalue",
                             title = comp_label,
                             subtitle = NULL,
                             caption = NULL,
                             pCutoff = p_cut,
                             FCcutoff = 2,
                             pointSize = 3.5,
                             labSize = 3,
                             typeConnectors = "closed",
                             drawConnectors = TRUE,
                             widthConnectors = 0.75,
                             colConnectors = 'grey30',
                             max.overlaps = 10)

    ggsave(file.path(plot_dir, paste0("volcano_", tumor, "_paired.pdf")),
           plot = plot, width = 8, height = 6)
  }
}
```

### Session Info

```{r session-info}
sessionInfo()
```
