---
title: "PCA visualization of miRNA expression across brain tumor histologies"
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Bicna Song
date: "2025-05-22"
---

This script generates PCA plots of miRNA expression data to evaluate batch effects and biological variation across samples.

It includes:
- PCA on uncorrected tumor samples
- PCA after batch correction using both removeBatchEffect() and ComBat_seq()
- PCA plots for each tumor type, including tumor-adjacent normals and disease-free controls (B1–B5), with labeled points for visualization

The analyses help assess whether histology or batch drives major variance in the dataset.

### Load required packages

```{r load packages, message=FALSE, warning=FALSE}
library(DESeq2)
library(tidyverse)
library(limma)
library(ggplot2)
library(rprojroot)
library(patchwork)
library(ggrepel)
library(sva)
library(matrixStats)
```

### Set directory paths

```{r dir paths}
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-mirna-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

source(file.path(root_dir, "figures", "theme.R"))
```

### Set file paths 

```{r file paths}
mirna_cts_file <- file.path(results_dir, "merged-miRNA-expression.tsv")
```

### Wrangle data

```{r wrangle data}
mirna_cts <- read_tsv(mirna_cts_file) %>%
  column_to_rownames("miRNA") %>%
  filter(rowSums(across(everything(), ~ . != 0)) >= 2)
```

### Load sample metadata df

```{r annotate-metadata}
# Fix annotation for specific sample
metadata_df <- read_delim(file.path(data_dir, "miRNA-sample-metadata.txt")) %>%
  mutate(
    sample_type = if_else(sample_id == "7316-7585-left-pons", "Normal-Tumor Adjacent", sample_type),
    histology = if_else(sample_id == "7316-7585-left-pons", "DIPG", histology)
  ) %>%
  mutate(histology = if_else(is.na(histology) | histology == "N/A", "Normal", histology)) %>%
  filter(sample_id %in% names(mirna_cts)) %>%
  arrange(match(sample_id, names(mirna_cts))) %>%
  mutate(batch = if_else(row_number() <= 5, "batch1", "batch2")) %>%
  mutate(condition = case_when(
    sample_type %in% c("Tumor", "Normal") ~ histology,
    sample_type == "Normal-Tumor Adjacent" ~ paste0(histology, "-adjacent Normal"),
    TRUE ~ NA_character_
  ))
```

### PCA on Tumor Samples Only (No Batch Correction)

```{r pca-tumor-only}
# Subset tumor samples
tumor_metadata <- metadata_df %>%
  filter(sample_type == "Tumor")

tumor_cts <- mirna_cts[, tumor_metadata$sample_id]

# Create DESeq2 object for tumor samples only
dds_tumor <- DESeqDataSetFromMatrix(countData = tumor_cts,
                                    colData = tumor_metadata,
                                    design = ~ batch + histology)

vsd_tumor <- varianceStabilizingTransformation(dds_tumor, blind = FALSE)

# Plot PCA
pca_tumor <- plotPCA(vsd_tumor, intgroup = c("histology", "batch"), returnData = TRUE)
percentVar <- round(100 * attr(pca_tumor, "percentVar"))

p1 <- ggplot(pca_tumor, aes(PC1, PC2, color = histology, shape = batch)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = histology), size = 3, max.overlaps = 50) +
  labs(title = "PCA: Tumor Samples Only (Uncorrected)",
       x = paste0("PC1: ", percentVar[1], "% variance"),
       y = paste0("PC2: ", percentVar[2], "% variance")) +
  theme_Publication()

p1
```

### PCA on Tumor Samples Only (Batch Correction)

```{r pca-corrected}
# Apply batch correction to vst data
corrected_counts <- removeBatchEffect(assay(vsd_tumor), batch = vsd_tumor$batch)

# Compute per-miRNA variance
miRNA_vars <- rowVars(corrected_counts)

# Pick the 500 miRNAs with highest variance
top500 <- order(miRNA_vars, decreasing = TRUE)[1:500]

# Subset the matrix to top 500 miRNAs
mat_top500 <- corrected_counts[top500, ]

# Run PCA on the transposed subset
pca_corrected <- prcomp(t(mat_top500), center = TRUE, scale. = FALSE)

# Prepare data
pca_df <- as.data.frame(pca_corrected$x)
pca_df$condition <- colData(vsd_tumor)$condition
pca_df$batch <- colData(vsd_tumor)$batch
pca_df$histology <- colData(vsd_tumor)$histology

percentVar_corrected <- round(100 * (pca_corrected$sdev^2 / sum(pca_corrected$sdev^2)))

p2 <- ggplot(pca_df, aes(PC1, PC2, color = condition, shape = batch)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = histology), size = 3, max.overlaps = 50) +
  labs(title = "PCA: Tumor Samples Only (Batch Corrected)",
       x = paste0("PC1: ", percentVar_corrected[1], "% variance"),
       y = paste0("PC2: ", percentVar_corrected[2], "% variance")) +
  theme_Publication()

p2
```

```{r}
p3 <- ggplot(pca_df, aes(PC2, PC3, color = condition, shape = batch)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = histology), size = 3, max.overlaps = 50) +
  labs(title = "PCA: Tumor Samples Only (Batch Corrected)",
       x = paste0("PC2: ", percentVar_corrected[2], "% variance"),
       y = paste0("PC3: ", percentVar_corrected[3], "% variance")) +
  theme_Publication()
p3
```

```{r}
p4 <- ggplot(pca_df, aes(PC3, PC4, color = condition, shape = batch)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = histology), size = 3, max.overlaps = 50) +
  labs(title = "PCA: Tumor Samples Only (Batch Corrected)",
       x = paste0("PC3: ", percentVar_corrected[3], "% variance"),
       y = paste0("PC4: ", percentVar_corrected[4], "% variance")) +
  theme_Publication()
p4
```

### PCA on Tumor Samples Only Using ComBat_seq

```{r pca-tumor-combat-seq}
# Apply ComBat_seq batch correction
combat_tumor <- ComBat_seq(as.matrix(tumor_cts), batch = tumor_metadata$batch)

# Create DESeq2 object and apply VST
dds_combat <- DESeqDataSetFromMatrix(countData = combat_tumor,
                                     colData = tumor_metadata,
                                     design = ~ histology)
vsd_combat <- varianceStabilizingTransformation(dds_combat, blind = FALSE)

# Perform PCA
pca_combat <- plotPCA(vsd_combat, intgroup = c("histology", "batch"), returnData = TRUE)
percentVar <- round(100 * attr(pca_combat, "percentVar"))

# Plot
p5 <- ggplot(pca_combat, aes(PC1, PC2, color = histology, shape = batch)) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = histology), size = 3, max.overlaps = 50) +
  labs(title = "PCA: Tumor Samples Only (ComBat Corrected)",
       x = paste0("PC1: ", percentVar[1], "% variance"),
       y = paste0("PC2: ", percentVar[2], "% variance")) +
  theme_Publication()

p5
```


### Save Plots

```{r save-plots}
ggsave(filename = file.path(plot_dir, "mirna-pca-tumor-only-uncorrected.pdf"), plot = p1, width = 10, height = 6)
ggsave(filename = file.path(plot_dir, "mirna-pca-tumor-only-corrected-PC1-PC2.pdf"), plot = p2, width = 10, height = 6)
ggsave(filename = file.path(plot_dir, "mirna-pca-tumor-only-corrected-PC2-PC3.pdf"), plot = p3, width = 10, height = 6)
ggsave(filename = file.path(plot_dir, "mirna-pca-tumor-only-corrected-PC3-PC4.pdf"), plot = p4, width = 10, height = 6)
ggsave(filename = file.path(plot_dir, "mirna-pca-tumor-only-corrected-combat.pdf"), plot = p5, width = 10, height = 6)
```


### PCA by Tumor Type vs Controls

```{r pca-each-tumor-vs-controls, message=FALSE, warning=FALSE}
tumor_types <- c("ATRT", "DIPG", "Ependymoma", "HGG", "LGG", "Medulloblastoma")
controls <- c("B1", "B2", "B3", "B4", "B5")

for (tumor in tumor_types) {
  tumor_samples <- metadata_df %>% filter(histology == tumor & sample_type == "Tumor")
  adjacent_normals <- metadata_df %>% filter(sample_type == "Normal-Tumor Adjacent" & histology == tumor)
  healthy_normals <- metadata_df %>% filter(sample_id %in% controls)

  subset_metadata <- bind_rows(tumor_samples, adjacent_normals, healthy_normals)
  subset_cts <- mirna_cts[, subset_metadata$sample_id]

  dds_subset <- DESeqDataSetFromMatrix(countData = subset_cts,
                                       colData = subset_metadata,
                                       design = ~ sample_type)
  vsd_subset <- varianceStabilizingTransformation(dds_subset, blind = FALSE)

  pca_data <- plotPCA(vsd_subset, intgroup = "sample_type", returnData = TRUE)
  percentVar <- round(100 * attr(pca_data, "percentVar"))
  pca_data$sample_id <- rownames(pca_data)

  p <- ggplot(pca_data, aes(PC1, PC2, color = sample_type, shape = sample_type)) +
    geom_point(size = 3) +
    geom_text_repel(aes(label = sample_id), size = 3, max.overlaps = 20) +
    labs(title = paste("PCA -", tumor, "vs Controls"),
         x = paste0("PC1: ", percentVar[1], "% variance"),
         y = paste0("PC2: ", percentVar[2], "% variance")) +
    theme_Publication()

  ggsave(filename = file.path(plot_dir, paste0("pca_", tumor, "_vs_controls.pdf")), plot = p, width = 10, height = 6)
}
```

### Session info

```{r session info}
sessionInfo()
```
