---
title: "Differential miRNA Expression Analysis"
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Bicna Song
date: "2025-05-23"
---

This script performs differential expression analysis of miRNA-seq data using DESeq2.

The goal is to identify differentially expressed miRNAs for each brain tumor histology by comparing:

- Tumor vs. Tumor-Adjacent Normal samples (e.g., DIPG vs DIPG-adjacent normal)
- Tumor vs. Disease-Free Normal brain samples (B1–B5)

### Load Required Packages

```{r load-packages, message=FALSE, warning=FALSE}
library(DESeq2)
library(tidyverse)
library(rprojroot)
library(readr)
library(EnhancedVolcano)
```

```{r}
install.packages("ashr") # this will be added to Dockerfile later.
library(ashr)
```

### Set Directories and Load Data

```{r setup-paths}
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-mirna-expression")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

source(file.path(root_dir, "figures", "theme.R"))
```

### Set file paths 

```{r file paths}
mirna_cts_file <- file.path(results_dir, "merged-miRNA-expression.tsv")
```

### Wrangle data

```{r wrangle data}
mirna_cts <- read_tsv(mirna_cts_file) %>%
  column_to_rownames("miRNA") %>%
  filter(rowSums(across(everything(), ~ . != 0)) >= 2)
```

### Load sample metadata df

```{r annotate-metadata}
# Fix annotation for specific sample
metadata_df <- read_delim(file.path(data_dir, "miRNA-sample-metadata.txt")) %>%
  mutate(
    sample_type = if_else(sample_id == "7316-7585-left-pons", "Normal-Tumor Adjacent", sample_type),
    histology = if_else(sample_id == "7316-7585-left-pons", "DIPG", histology)
  ) %>%
  mutate(histology = if_else(is.na(histology) | histology == "N/A", "Normal", histology)) %>%
  filter(sample_id %in% names(mirna_cts)) %>%
  arrange(match(sample_id, names(mirna_cts))) %>%
  mutate(batch = if_else(row_number() <= 5, "batch1", "batch2")) %>%
  mutate(condition = case_when(
    sample_type %in% c("Tumor", "Normal") ~ histology,
    sample_type == "Normal-Tumor Adjacent" ~ paste0(histology, "-adjacent Normal"),
    TRUE ~ NA_character_
  ))

stopifnot(identical(metadata_df$sample_id, colnames(mirna_cts)))
```

### Perform Differential Expression for Each Histology

```{r run-de-by-histology, message=FALSE, warning=FALSE}
normal_ids <- c("B1", "B2", "B3", "B4", "B5")
histologies <- c("ATRT", "DIPG", "Ependymoma", "HGG", "Medulloblastoma")

for (tumor in histologies) {
  adj_label <- paste0(tumor, "-adjacent Normal")

  # Subset metadata
  meta_adj <- metadata_df %>% filter(condition %in% c(tumor, adj_label))
  meta_healthy <- metadata_df %>% filter(sample_id %in% c(normal_ids, sample_id[condition == tumor])) %>% 
    filter(condition %in% c(tumor, "Normal"))

  # Select count data
  counts_adj <- mirna_cts[, meta_adj$sample_id]
  counts_healthy <- mirna_cts[, meta_healthy$sample_id]

  # Tumor vs Adjacent Normal
  design_adj <- if (tumor == "DIPG") ~ batch + condition else ~ condition
  dds_adj <- DESeqDataSetFromMatrix(countData = counts_adj, colData = meta_adj, design = design_adj)
  dds_adj$condition <- droplevels(dds_adj$condition)
  dds_adj <- DESeq(dds_adj)
  res_adj <- results(dds_adj, contrast = c("condition", tumor, adj_label))
  res_adj <- lfcShrink(dds_adj, contrast = c("condition", tumor, adj_label), res = res_adj, type = "ashr")
  write_csv(as.data.frame(res_adj) %>% rownames_to_column("miRNA"),
          file.path(results_dir, paste0("DESeq2_", tumor, "_vs_adjNormal.csv")))


  # Tumor vs Healthy Normal
  design_healthy <- if (tumor == "DIPG") ~ batch + condition else ~ condition
  dds_healthy <- DESeqDataSetFromMatrix(countData = counts_healthy, colData = meta_healthy, design = design_healthy)
  dds_healthy$condition <- droplevels(dds_healthy$condition)
  dds_healthy <- DESeq(dds_healthy)
  res_healthy <- results(dds_healthy, contrast = c("condition", tumor, "Normal"))
  res_healthy <- lfcShrink(dds_healthy, contrast = c("condition", tumor, "Normal"), res = res_healthy, type = "ashr")
  write_csv(as.data.frame(res_healthy) %>% rownames_to_column("miRNA"),
          file.path(results_dir, paste0("DESeq2_", tumor, "_vs_healthyNormal.csv")))
}
```

### Summary Table: Number of Up/Down-regulated miRNAs

```{r summary-table}
summary_df <- tibble()

for (tumor in histologies) {
  for (type in c("adjNormal", "healthyNormal")) {
    file_path <- file.path(results_dir, paste0("DESeq2_", tumor, "_vs_", type, ".csv"))
    if (file.exists(file_path)) {
      res_df <- read_csv(file_path)
      res_df <- res_df %>%filter(!is.na(padj) & padj < 0.05)
      write_csv(res_df, file.path(results_dir, paste0("DESeq2_", tumor, "_vs_", type, "_sig.csv"))) # Save Significant DE Results Separately
      up <- sum(res_df$log2FoldChange > 0)
      down <- sum(res_df$log2FoldChange < 0)
      summary_df <- bind_rows(summary_df, 
        tibble(Comparison = paste0(tumor, " vs ", type), 
               Upregulated = up, 
               Downregulated = down))
    }
  }
}

summary_df
```

```{r}
# Save to CSV
write_csv(summary_df, file.path(results_dir, "DE_summary_counts.csv"))
```

### Volcano Plots for DE Comparisons

```{r volcano-plots, message=FALSE, warning=FALSE}
for (tumor in histologies) {
  for (type in c("adjNormal", "healthyNormal")) {
    file_path <- file.path(results_dir, paste0("DESeq2_", tumor, "_vs_", type, ".csv"))
    if (file.exists(file_path)) {
      res_df <- read_csv(file_path)
      rownames(res_df) <- res_df$miRNA %||% rownames(res_df)  # Fallback for rownames if included
      p_cut <- max(res_df$pvalue[res_df$padj < 0.05], na.rm = TRUE)
      comp_label <- paste(tumor, "vs", gsub("Normal", "normal", type))
      
      plot <- EnhancedVolcano(res_df,
                              lab = rownames(res_df),
                              x = "log2FoldChange",
                              y = "pvalue",
                              title = comp_label,
                              subtitle = NULL,
                              caption = NULL,
                              pCutoff = p_cut,
                              FCcutoff = 2,
                              pointSize = 3.5,
                              labSize = 3,
                              typeConnectors = "closed",
                              drawConnectors = TRUE,
                              widthConnectors = 0.75,
                              colConnectors = 'grey30',
                              max.overlaps = 10)

      ggsave(file.path(plot_dir, paste0("volcano_", tumor, "_vs_", type, ".pdf")),
             plot = plot, width = 8, height = 6)
    }
  }
}
```

### Session Info

```{r session-info}
sessionInfo()
```
