---
title: 'Human miRNA-RNA interactions in DIPG/DMG'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2025"
---

This script identifies coordinated alterations in miRNA and target mRNA abundance in DIPG/DMG versus normal pons

Load packages

```{r load packages}
library(tidyverse)
library(ggplot2)
```

Set directory paths

```{r dir paths}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "human-mirna-rna-interactions")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results")
```

Set file paths 

```{r file paths}
de_mirna_file <- file.path(root_dir, "analyses",
                           "human-mirna-differential-expression",
                           "results",
                           "mirna-differential-expr-deseq2-dipg-dmg-versus-normal.tsv")

de_rna_file <- file.path(root_dir, "analyses",
                           "human-rna-differential-expression",
                           "results",
                           "mrna-differential-expression-deseq2-dipg-dmg-versus-normal.tsv")

targets_file <- file.path(data_dir,
                          "miRNA_Target_anno.csv")

mirTarBase_SE_R_file <- file.path(input_dir, "miRTarBase_SE_R.csv")
mirTarBase_SE_W_file <- file.path(input_dir, "miRTarBase_SE_W.csv")
mirTarBase_SE_WR_file <- file.path(input_dir, "miRTarBase_SE_WR.csv")
mirTarBase_WE_Clip_file <- file.path(input_dir, "miRTarBase_WE_Clip.tsv.gz")
mirTarBase_WE_Other_file <- file.path(input_dir, "miRTarBase_WE_Other.csv")
```

Wrangle data

```{r wrangle data}
de_mirna_df <- read_tsv(de_mirna_file)

de_rna_df <- read_tsv(de_rna_file)

targets_df <- read_csv(targets_file) %>%
  # keep only unique miRNA-gene id pairs
  distinct(Seq1, GeneID, .keep_all = TRUE)

mirTarBase_df <- read_csv(mirTarBase_SE_R_file) %>%
  bind_rows(read_csv(mirTarBase_SE_W_file)) %>%
  bind_rows(read_csv(mirTarBase_SE_WR_file)) %>%
  bind_rows(read_csv(mirTarBase_WE_Other_file)) %>%
  bind_rows(read_tsv(mirTarBase_WE_Clip_file)) %>%
  dplyr::filter(grepl("hsa", miRNA))
```

Save merged miRTarBase human miRNA predicted target df

```{r}
write_tsv(mirTarBase_df,
          file.path(results_dir, "human-mirna-target-predictions-miRTarBase.tsv.gz"))
```

For annotated DE miRNAs: add predicted target genes from miRTarBase and DE results

```{r}
known_mirna_target_df <- de_mirna_df %>%
  dplyr::filter(padj < 0.1) %>%
  left_join(mirTarBase_df %>% dplyr::select(miRNA,
                                         `Target Gene`,
                                         `Support Type`,
                                         Experiments),
            by = c("mirna_id" = "miRNA")) %>%
  dplyr::rename(target_gene_id = `Target Gene`) %>%
  left_join(de_rna_df,
            by = c("target_gene_id" = "gene_name"),
            suffix = c("_miRNA", "_target_gene")) %>%
  dplyr::filter(!is.na(baseMean_target_gene),
                !is.na(target_gene_id)) %>%
  distinct(mirna_id, target_gene_id, .keep_all = TRUE)
```

For annotated DE miRNAs: add predicted target genes from Azenta target gene prediction, and DE results

```{r}
novel_mirna_target_df <- de_mirna_df %>%
  dplyr::filter(padj < 0.1,
                grepl("Novel", mirna_id)) %>%
  left_join(targets_df %>% dplyr::select(Seq1,
                                         GeneID),
            by = c("mirna_id" = "Seq1")) %>%
  dplyr::rename(target_gene_id = GeneID) %>%
  left_join(de_rna_df,
            by = c("target_gene_id" = "gene_id"),
            suffix = c("_miRNA", "_target_gene")) %>%
  dplyr::filter(!is.na(baseMean_target_gene),
                !is.na(target_gene_id)) %>%
  distinct(mirna_id, target_gene_id, .keep_all = TRUE)
```

Save outputs

```{r save output}
write_tsv(known_mirna_target_df,
          file.path(results_dir,
                    "known-mirna-target-gene-differential-expr-dipg-dmg-versus-normal.tsv"))

write_tsv(novel_mirna_target_df,
          file.path(results_dir,
                    "novel-mirna-target-gene-differential-expr-dipg-dmg-versus-normal.tsv"))
```

Print session info

```{r session info}
sessionInfo()
```